---
title: "T50 hot and cold"
format: html
editor: visual
---

### **LOADING LIBRARY**

```{r}
#| label: loading library
#| message: false

library(tidyverse)
library(readxl)
library(here)
library(thermalTolerance) 

# remember to re-install if Richard changes package, found on github, installed using: 
# DO NOT install the most recent version. remotes::install_github("richardjtelford/thermalTolerance") 
# this package estimates of 3 different PSII heat tolerance metrics, Tcrit, T50 and T95. Can take each individual nr plant into account, fits curves by randomly resampling for each individual by bootstrapping.
```

### CLEAN DATA IMPORT

```{r}
#| label: data import  
#| message: false

#data import of cleaned data

clean_data <- read.csv(here("clean_data", "DURIN_Clean_Seasonal_Thermal_Tolerance_Bente_Msc.csv"))

view(clean_data) # check import 
```

#### Levels - need to change manes of age classes at some point!

```{r}
#| label: levels 
#| message: false

# setting levels for figures
clean_data$age_class <- factor((clean_data$age_class),
                         levels = c("Newly Burnt", "Pioneer", 
                                    "Building", "Mature"))
clean_data$month <- factor((clean_data$month),
                     levels = c("November", "January"))

```

### HOT AND COLD SUBSETS

```{r}
#| label: subsetting  
#| message: false

# making hot subset all clean data
hot_subset <- clean_data |> 
  filter(proj_2025 == "TT Hot") |> # filter for TT Hot
  filter(!is.na(fv_fm_final)) |>  # removing row with NA for fv_fm_final
   select(month, age_class, plant_nr, temp_exp, fv_fm_final, unique_plant_id)


# making cold subset all clean data
cold_subset <- clean_data |> 
  filter(proj_2025 == "TT Cold")|> # filter for TT Cold 
  filter(!is.na(fv_fm_final)) |>  # removing row with NA for fv_fm_final
   select(month, age_class, plant_nr, temp_exp, fv_fm_final, unique_plant_id)

# filter further if certain datapoints or whole individuals should be taken out. 
```

## HOT

### Raw data scatter plot

Making scatter plot for visualization. Points for each individual plant represented by colour and connected by lines. Faceted by month x age class.

```{r}
#| label: hot visualization plot  
#| message: false
   
# summary vizualisation plot for hot treatment faceted by month x age class 
ggplot(hot_subset, aes(x = as.numeric(as.character(temp_exp)), y = fv_fm_final, 
                       colour = factor(plant_nr))) +  
  
  geom_line(alpha = 0.4) +  # one line per plant
  geom_point(alpha = 0.6, size = 1.5) + # one point per measurement
  facet_grid(month ~ age_class) + # facet by month x age class

# add labels  
  labs( 
    title = "Fv/Fm response across heat treatments from 20 C to 56 C",
    subtitle = "Each colour and line represent one individual plant. ", 
    x = "Treatment temperature (C)",
    y = "Fv/Fm",
    colour = "Plant nr"
  ) +

#set theme  
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) 

# saving plot to output folder
ggsave(here("output", "hot_scatter_line_plot.png"), 
       width = 10, height = 6, dpi = 300) 
```

Problematic single points:

-   January_Pioneer_1 (low Fv/Fm at 20 C control)

-   January_Building_3 (low Fv/Fm at 20 C control - although whole curve bounce up and down)

-   January_Pioneer_6 (Fv/Fm increase again, is high at 56 C)

Problematic curves on whole individual level (curves that bounce up and down):

-   November_Newly Burnt_1, November_Pioneer_2, November_Building_2, November_Building_7, January_Building_4, January_Building_2, January_Mature_3, January_Mature_4, January_Pioneer_2, January_Pioneer_5, January_Pioneer_4, January_Pioneer_3, January_Building_6

Most November individuals have a steep drop in Fv/Fm from -10 C to - 20 C.

Most January individuals do not have a steep drop in Fv/Fm at more negative temperatures, although with a few exceptions. Indicate greater freezing tolerance in January compared to November. Biologically meaningful, statistically harder to use.

January_Building_4 has an odd low point (0.00 = dead) at -15 before coming back to life at -20. Checked with after-treatment photos, leaves look healthy but one is missing. Therefore less biomass for measuring Fv/Fm than ususal, however should not end up with 0.00. Unlikely to be representative value for -15 treatment for this month x age class combination. Check what happens to estimate if set to NA.

#### Making remark column for odd hot points/curves

```{r}

# NOT WORKING YET 

hot_subset$hot_remark <- NA_character_

# adding remarks for specific individuals 
# need to move down , stich on afterwards?
hot_subset |> 
  mutate(
  hot_remark = case_when(
      new_id == "January_Pioneer_1" ~ "; low starting value ; bouncy curve",
      id == "January_Building_3" ~ "; low starting value",
      id == "November_Newly Burnt_1" ~ "; bounces back once",
      id == "November_Pioneer_2" ~ "; bounces back once",
      id == "November_Building_2" ~ "; bounces back once",
      id == "November_Building_7" ~ "; bounces back once",
      id == "January_Building_4" ~ "; low starting value ; bounces back once",
      id == "January_Building_2" ~ "; bouncy curve",
      id == "January_Mature_3" ~ "; bouncy curve",
      id == "January_Mature_4" ~ "; bouncy curve",
      id == "January_Pioneer_2" ~ "; low starting value",
      id == "January_Pioneer_5" ~ "; low starting value",
      id == "January_Pioneer_4" ~ "; bouncy curve",
      id == "January_Pioneer_3" ~ "; bouncy curve",
      id == "January_Building_6" ~ "; bounces back once",
      id == "January_Pioneer_6" ~ "; high ending value",
      TRUE ~ remark))
```

### Testing psiiht function on individual level hot with 50 boots

Estimating curves using psiiht function. The function fits a thermal tolerance curve based on Fv/Fm vs. temperature. Can choose to run grouped (month x age class) or on individual plant level. Will run on individual plant level here.

Remember to set.seed before every new psiiht, so the random number generator for bootstrapping start at the same point each time. So that results of bootstrapping is consistent if rerunning script.

boots = n control how many (n) bootstrap iterations, ie. how many times a curve is fit based on random re-sampeling from the data set. Returns 95% CI for the curves, not T50 directly. Keep boots low for testing, high \>500 for analysis.

When copied in from github, "htdata" is demo data from Perez 2020, must modify to fit own dataset before use

#### Exploratory: CONTROL TEMP 20C

```{r}
#| label: psiiht for hot 
#| message: false

# stich columns together, using unite
united_columns_hot <- unite(
  hot_subset,
  col = "new_id", 
  month, age_class, plant_nr, # fitting nls for each individual plant
  sep = "_",
  remove = TRUE
)

set.seed(8) # set seed before every psiiht  
# psiiht function for fitting thermal tolerance curve 
htol_20_all <- psiiht(temperature = united_columns_hot$temp_exp, 
               warming = TRUE,
               fvfm = united_columns_hot$fv_fm_final, 
               control_temp = 20,
               id = united_columns_hot$new_id,
               boots = 50 ) #increase in final

htol_summary_20_all <- summary(htol_20_all) |> 
          select(-Tcrit.mn, -T95.mn) # see summary output, only interested in T50

autoplot(htol_20_all) # plots logistic curves and dashed lines for Tcrit, T50, T95 


#extracting the CI for the individual T50 found from bootstrapping
htol_20_all_t50_ci <- purrr::map_dfr(names(htol_20_all$results), function(id) {
  t50_boots <- htol_20_all$results[[id]]$T50
  data.frame(
    id    = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = quantile(t50_boots, 0.025, na.rm = TRUE),
    T50.hi  = quantile(t50_boots, 0.975, na.rm = TRUE)
  )
})

# get a warning, this is because the psiiht could not fit a T50 for all individuals

```

Run with 50 boots for testing function with all clean hot data. Returns relatively high prop_success for most individuals, with some exeptions:

Can not fit curves (prop_success = NA) for two individuals; January_Pioneer_1 and January\_ Building_3 -This is the same individuals that had abnormally low starting Fv/Fm at 20 C control. Will try to run the model on 38 C as control temperature to see if these two individuals can be included as well.

Returns very low success rates (prop_success \<0.5) for January_Building_4

Returns low success rates (prop_success \<0.8) for Building_January_2, January_Mature_4, November_Pioneer_1, January_Newly Burnt_4, January_Poineer_4, January_Mature_6, November_Pioneer_5 and November_Building_2

#### Exploratory: CONTROL TEMP 38C

Taking into account that January_Pioneer_1 and January_Building_3 have very low starting values, trying to run the function with control temp 38 to see if those individuals can be included

```{r}
#| label: psiiht for hot from 38
#| message: false

set.seed(8) # set seed before every psiiht  
# psiiht function for fitting thermal tolerance curve starting at 38 C 
htol_38 <- psiiht(temperature = united_columns_hot$temp_exp, 
               warming = TRUE,
               fvfm = united_columns_hot$fv_fm_final, 
               control_temp = 38, #NB! control changed to 38
               id = united_columns_hot$new_id,
               boots = 50 ) #increase in final

htol_summary_38 <- summary(htol_38) |> 
      select(-Tcrit.mn, -T95.mn) # see summary output, only interested in T50

autoplot(htol_38) # plots logistic curves and dashed lines for Tcrit, T50, T95 

#extracting the CI for the individual T50 found from bootstrapping
htol_38_t50_ci <- purrr::map_dfr(names(htol_38$results), function(id) {
  t50_boots <- htol_38$results[[id]]$T50
  data.frame(
    id    = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = quantile(t50_boots, 0.025, na.rm = TRUE),
    T50.hi  = quantile(t50_boots, 0.975, na.rm = TRUE)
  )
})
```

Still cannot fit T50s for the two problematic individuals.

#### COMPARING MODEL WITH CONTROL = 20 VS. CONTROL = 38 ind + group level

Comparing the T50 mean values on individual level. To be able to see if specific plants act differently between the two models, ie. if the models agree.

```{r}
#| label: 20 vs 38 individual 
#| message: false

# create comparison object 
comparison_20_vs_38 <- htol_20_all_t50_ci |>
  rename(T50_20 = T50.mn, low_20 = T50.low, hi_20 = T50.hi) |>
  inner_join(
    htol_38_t50_ci |>
      rename(T50_38 = T50.mn, low_38 = T50.low, hi_38 = T50.hi),
    by = "id"
  ) |>
  separate(id, into = c("month", "age_class", "plant_nr"), sep = "_", remove = FALSE) |>
  mutate(
    T50_diff   = T50_38 - T50_20,
    ci_overlap = !(hi_20 < low_38 | hi_38 < low_20)
  ) |>
  select(id, month, age_class, plant_nr, T50_20, T50_38, T50_diff, ci_overlap)


# plot visually comparing T50 results between 20_all and 38 models with CI
htol_compare_20_vs_38_ind <- bind_rows(
  htol_20_all_t50_ci |> mutate(run = "control_20"),
  htol_38_t50_ci     |> mutate(run = "control_38")
) |>
  ggplot(aes(x = T50.mn, y = id, color = run, shape = run)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbarh(aes(xmin = T50.low, xmax = T50.hi),
                 height = 0.2,
                 position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("control_20" = "grey50", "control_38" = "steelblue")) +
  labs(x = "T50 (C)", y = NULL, title = "T50 comparison: control 20 vs 38",
       color = NULL, shape = NULL) +
  theme_bw()

print(htol_compare_20_vs_38_ind)

# saving plot to output folder individual
ggsave(here("output", "htol_plot_compare_20_vs_38_ind.png"), 
       width = 10, height = 6, dpi = 300) 



# individual level comparison 
comparison_20_vs_38 |>
  select(month, age_class, plant_nr, T50_20, T50_38, T50_diff, ci_overlap) |>
  arrange(month, age_class, plant_nr) |>
  print(n = Inf)


# group level paired t-test per month and age class 
comparison_20_vs_38 |>
  group_by(month, age_class) |>
  summarise(
    t_test = list(t.test(T50_20, T50_38, paired = TRUE)),
    .groups = "drop"
  ) |>
  mutate(
    p_value   = purrr::map_dbl(t_test, ~ .x$p.value),
    mean_diff = purrr::map_dbl(t_test, ~ .x$estimate)
  ) |>
  select(month, age_class, p_value, mean_diff)


# compare prop_success between models 
prop_compare_20_vs_38_ind <- bind_rows(
  htol_summary_20_all |> mutate(model = "control 20"),
  htol_summary_38     |> mutate(model = "control 38")
) |>
  select(id, prop_success, model) |>
  pivot_wider(names_from = model, values_from = prop_success)

prop_compare_20_vs_38_ind  # equal for both models

```

Individual level: Big differences in predicted T50 (over 1 C diff) for: January_Mature_4 (2.1 C), January_Mature_7 (1.2 C), January_Pioneer_4 (1.1 C), January_Pioneer_5 (-1.5 C), January_Pioneer_6 (-1.0)

For most groups the two models produce statistically interchangeable T50 estimates, because p-value \> 0.05.

But January Building shows a significant difference (p = 0.037, mean difference = 0.26C). Probably because of lower and more variable starting Fv/Fm in January individuals. Would make the model more sensitive to control temperature choice.

Also, for November mature the p = 0.02, but difference in mean T50 is only 0.06, so not very meaningful.

How does this affect choice of control temp? Which is the better option? Need to discuss.

NB: CANNOT SEE THE LAST 4 PIONEERS IN THE SUMMARY - WHY?

#### Exploratory: comparing T50s and prop_success when setting odd points to NA

#### NOVEMBER SUBSET CLEANED AND RAW

Starting with just the November plant, as the data here is more tidy. Setting two specific data points to NA, due to four individuals2 bouncing back to life after fvfm decline at erlier temps. To see if it changes the predicted T50 values both on the individual plant level, and on the group level.

```{r}
#| label: november raw and clean 
#| message: false

# subset to November
subset_nov <- hot_subset |>
  filter(month == "November")

# setting four specific data pions which had unlikely bounce_back in fvfm after initial decline as NA
subset_nov_cleaned <- subset_nov |>
  mutate(fv_fm_final = case_when(
    age_class == "Newly Burnt" & plant_nr == 1 & temp_exp == 50 ~ NA_real_,
    age_class == "Pioneer"     & plant_nr == 2 & temp_exp == 50 ~ NA_real_,
    age_class == "Building"    & plant_nr == 2 & temp_exp == 50 ~ NA_real_,
    age_class == "Building"    & plant_nr == 7 & temp_exp == 50 ~ NA_real_,
    TRUE ~ fv_fm_final
  ))

# fit models for each
# fit model for all nov data
united_nov_raw <- unite(subset_nov, col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

set.seed(8)
htol_nov_raw <- psiiht(temperature = united_nov_raw$temp_exp, 
                       warming = TRUE, 
                       fvfm = united_nov_raw$fv_fm_final, 
                       control_temp = 20,
                       id = united_nov_raw$new_id,
                       boots = 50)

# fit model for cleaned nov data
united_nov_cleaned <- unite(subset_nov_cleaned, col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

set.seed(8)
htol_nov_cleaned <- psiiht(temperature = united_nov_cleaned$temp_exp, 
                           warming = TRUE,
                           fvfm = united_nov_cleaned$fv_fm_final,
                           control_temp = 20,
                           id = united_nov_cleaned$new_id,
                           boots = 50)


# extract bootstrap CI for individual T50 for each

# CI extraction for nov raw data
nov_raw_t50_ci <- purrr::map_dfr(names(htol_nov_raw$results), function(id) {
  t50_boots <- htol_nov_raw$results[[id]]$T50
  if (!is.numeric(t50_boots) | length(t50_boots) == 0) {
    return(data.frame(id = id, T50.mn = NA, T50.low = NA, T50.hi = NA))
  }
  data.frame(
    id      = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = unname(quantile(t50_boots, 0.025, na.rm = TRUE)),
    T50.hi  = unname(quantile(t50_boots, 0.975, na.rm = TRUE))
  )
})

# CI extraction for nov clean data 
nov_cleaned_t50_ci <- purrr::map_dfr(names(htol_nov_cleaned$results), function(id) {
  t50_boots <- htol_nov_cleaned$results[[id]]$T50
  if (!is.numeric(t50_boots) | length(t50_boots) == 0) {
    return(data.frame(id = id, T50.mn = NA, T50.low = NA, T50.hi = NA))
  }
  data.frame(
    id      = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = unname(quantile(t50_boots, 0.025, na.rm = TRUE)),
    T50.hi  = unname(quantile(t50_boots, 0.975, na.rm = TRUE))
  )
})

```

Four odd Fv/Fm values representing bounce-backs after initial decline were identified and set to NA for testing model.

#### COMPARING MODEL FOR RAW VS CLEAN NOVEMBER DATA

```{r}
#| label: compare november raw vs clean 
#| message: false

# making comparison object 
nov_comparison <- nov_raw_t50_ci |>
  rename(T50_raw = T50.mn, raw.low = T50.low, raw.hi = T50.hi) |>
  inner_join(
    nov_cleaned_t50_ci |>
      rename(T50_cleaned = T50.mn, cleaned.low = T50.low, cleaned.hi = T50.hi),
    by = "id"
  ) |>
  separate(id, into = c("month", "age_class", "plant_nr"), sep = "_", remove = FALSE) |>
  mutate(
    T50_diff   = T50_cleaned - T50_raw,
    ci_overlap = !(raw.hi < cleaned.low | cleaned.hi < raw.low)
  ) |>
  select(id, month, age_class, plant_nr, T50_raw, T50_cleaned, T50_diff, ci_overlap)


# plot visual comparison with CI individual
htol_compare_nov_raw_vs_cleaned <- bind_rows(
  nov_raw_t50_ci     |> mutate(version = "raw"),
  nov_cleaned_t50_ci |> mutate(version = "cleaned")
) |>
  ggplot(aes(x = T50.mn, y = id, color = version, shape = version)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_errorbarh(aes(xmin = T50.low, xmax = T50.hi),
                 height = 0.2,
                 position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("raw" = "grey50", "cleaned" = "steelblue")) +
  labs(x = "T50 (C)", y = NULL, title = "November T50: raw vs cleaned",
       color = NULL, shape = NULL) +
  theme_bw()

print(htol_compare_nov_raw_vs_cleaned)

# save plot 
ggsave(here("output", "htol_compare_nov_raw_vs_cleaned.png"),
       width = 10, height = 6, dpi = 300)

# looks like CIs are overlapping also for the individual where NA has been introduced. 


# individual level comparison all age classes and months 
nov_comparison |>
  select(month, age_class, plant_nr, T50_raw, T50_cleaned, T50_diff, ci_overlap) |>
  arrange(month, age_class, plant_nr) |>
  print(n = Inf)

# group level paired t-test per age class
nov_comparison |>
  group_by(age_class) |>
  summarise(t_test = list(t.test(T50_raw, T50_cleaned, paired = TRUE))) |>
  mutate(
    p_value   = purrr::map_dbl(t_test, ~ .x$p.value),
    mean_diff = purrr::map_dbl(t_test, ~ .x$estimate)
  ) |>
  select(age_class, p_value, mean_diff)


# compare prop_success between raw and cleaned november models
prop_compare_nov_raw_vs_cleaned <- bind_rows(
  summary(htol_nov_raw)     |> mutate(model = "raw"),
  summary(htol_nov_cleaned) |> mutate(model = "cleaned")
) |>
  select(id, prop_success, model) |>
  pivot_wider(names_from = model, values_from = prop_success)

prop_compare_nov_raw_vs_cleaned
```

Individual level: High differences in predicted T50 (over 1C) for: November_Pioneer_2 (-1.4 C)

Paired t-tests comparing T50 estimates from raw and cleaned datasets found no significant differences for any age class (grouped) (all p \> 0.05, mean differences \< 0.12B0C), indicating that the cleaning had small effect on the results.

So points could be set as NA without big differences in T50 results on group level.

Prop_success stays relatively similar - except for November_Building_2 which is reduced from 0.84 to 0.5. Likely because there is less points to choose from after setting NA.

#### JANUARY SUBSET CLEANED COMPARED TO RAW

do the same as above for january data. more weird points that needs adressing here. maybe better to run on subsets instead of all january data at once

```{r}
#| label: january raw vs cleaned   
#| message: false

# subset to January
subset_jan <- hot_subset |>
  filter(month == "January")

# setting specific points with abnormal starting and ending FvFm values to NA to se how it affects the T50 predictions 
subset_jan_cleaned <- subset_jan |>
  mutate(fv_fm_final = case_when(
    age_class == "Pioneer"  & plant_nr == 1 & temp_exp == 20 ~ NA_real_,  # low starting value
    age_class == "Pioneer"  & plant_nr == 2 & temp_exp == 20 ~ NA_real_,  # low starting value
    age_class == "Pioneer"  & plant_nr == 3 & temp_exp == 20 ~ NA_real_,  # low curve
    age_class == "Pioneer"  & plant_nr == 4 & temp_exp == 20 ~ NA_real_,  # low curve
    age_class == "Pioneer"  & plant_nr == 5 & temp_exp == 20 ~ NA_real_,  # low starting value
    age_class == "Pioneer"  & plant_nr == 6 & temp_exp == 56 ~ NA_real_,  # high ending value
    age_class == "Building" & plant_nr == 3 & temp_exp == 20 ~ NA_real_,  # low starting value
    age_class == "Building" & plant_nr == 4 & temp_exp == 20 ~ NA_real_,  # low starting value
    TRUE ~ fv_fm_final
  ))

# verify points removed
subset_jan_cleaned |>
  group_by(age_class, plant_nr) |>
  summarise(removed = sum(is.na(fv_fm_final)), .groups = "drop") |>
  filter(removed > 0)

# fit models for each
# fit model for all jan data
united_jan_raw <- unite(subset_jan, col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

set.seed(8)
htol_jan_raw <- psiiht(temperature = united_jan_raw$temp_exp, 
                       warming = TRUE, 
                       fvfm = united_jan_raw$fv_fm_final, 
                       control_temp = 20,
                       id = united_jan_raw$new_id,
                       boots = 50)

# fit model for cleaned nov data
united_jan_cleaned <- unite(subset_jan_cleaned, col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

set.seed(8)
htol_jan_cleaned <- psiiht(temperature = united_jan_cleaned$temp_exp, 
                           warming = TRUE,
                           fvfm = united_jan_cleaned$fv_fm_final,
                           control_temp = 20,
                           id = united_jan_cleaned$new_id,
                           boots = 50)


# extract bootstrap CI for individual T50 for each

# CI extraction for nov raw data
jan_raw_t50_ci <- purrr::map_dfr(names(htol_jan_raw$results), function(id) {
  t50_boots <- htol_jan_raw$results[[id]]$T50
  if (!is.numeric(t50_boots) | length(t50_boots) == 0) {
    return(data.frame(id = id, T50.mn = NA, T50.low = NA, T50.hi = NA))
  }
  data.frame(
    id      = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = unname(quantile(t50_boots, 0.025, na.rm = TRUE)),
    T50.hi  = unname(quantile(t50_boots, 0.975, na.rm = TRUE))
  )
})

# CI extraction for nov clean data 
jan_cleaned_t50_ci <- purrr::map_dfr(names(htol_nov_cleaned$results), function(id) {
  t50_boots <- htol_jan_cleaned$results[[id]]$T50
  if (!is.numeric(t50_boots) | length(t50_boots) == 0) {
    return(data.frame(id = id, T50.mn = NA, T50.low = NA, T50.hi = NA))
  }
  data.frame(
    id      = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = unname(quantile(t50_boots, 0.025, na.rm = TRUE)),
    T50.hi  = unname(quantile(t50_boots, 0.975, na.rm = TRUE))
  )
})
```

MODEL FAILS WHEN CONTROL TEMPERATURE IS SET TO NA. Because the T50 is calculated based on the starting value at control. If control is missing, it cannot calculate T50.

#### EXPLORATORY: COMPARISON OF DATA AFTER TALKING TO RICHARD 16/4

After discussion, the only datapoint worth setting to NA is January_Pioneer_1, 20 C, as this seems unlikely. All other points could be correct, impossible to know which ones are incorrect. Therefore keep them in.

Set January_Pioneer_1, 20 C to NA and do sensitivity analysis. Not reccommented to use t-test afterall.

```{r}
#| label: model full data minus jan_pio_1   
#| message: false

# subset to clean hot data
# setting January_Pioneer_1, 20 C data point to NA, as this is the only value that very certanly is incorrect
# update: control value CANNOT be set to NA, as the model needs a starting value. 
# removing whole individual instead. 

# does not work when settin the point to NA, therefore removing whole individual instead

hot_cleaned <- hot_subset |>
  filter(!(age_class == "Pioneer" & month == "January" & plant_nr == 1))

# fit model for cleaned data
united_hot_cleaned <- unite(hot_cleaned, 
                            col = "new_id", month, age_class, plant_nr,
                            sep = "_", remove = TRUE)

set.seed(8)
htol_20_cleaned <- psiiht(temperature = united_hot_cleaned$temp_exp, 
                       warming = TRUE, 
                       fvfm = united_hot_cleaned$fv_fm_final, 
                       control_temp = 20,
                       id = united_hot_cleaned$new_id,
                       boots = 50)

htol_summary_20_cleaned <- summary(htol_20_cleaned) |> 
          select(-Tcrit.mn, -T95.mn) # see summary output, only interested in T50

autoplot(htol_20_cleaned) # plots logistic curves and dashed lines for Tcrit, T50, T95 

#extracting the CI for the individual T50 found from bootstrapping
htol_20_clean_t50_ci <- purrr::map_dfr(names(htol_20_cleaned$results), function(id) {
  t50_boots <- htol_20_cleaned$results[[id]]$T50
  data.frame(
    id    = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = quantile(t50_boots, 0.025, na.rm = TRUE),
    T50.hi  = quantile(t50_boots, 0.975, na.rm = TRUE)
  )
})

# get a warning because some individuals have NA instead of means
```

Realized that January_Pioneer_1 was actually one of the two individuals the model failed to calculate T50 for, so removing this plant explicitly does not affect the group mean T50 whatsoever.

### 500 BOOTS HOT PSIIHT FOR FINAL T50:

Based on findings and conversation with Richard about only one point - that does not affect the mean - could be justified to set to NA, I choose to run the model for T50 hot with all hot Fv/Fm data included.

```{r}
#| label: hot 500 boots
#| message: false

# run psiiht with 500 boots on full hot_subset

# stich columns together, using unite, same as forfirst test run
united_columns_hot <- unite(
  hot_subset,
  col = "new_id", 
  month, age_class, plant_nr, # fitting nls for each individual plant
  sep = "_",
  remove = TRUE
)
set.seed(8) # set seed before every psiiht  
# psiiht function for fitting thermal tolerance curve 
htol_500_boots <- psiiht(temperature = united_columns_hot$temp_exp, 
               warming = TRUE,
               fvfm = united_columns_hot$fv_fm_final, 
               control_temp = 20,
               id = united_columns_hot$new_id,
               boots = 500 ) #increase to 500 in final

htol_summary_500_boots <- summary(htol_500_boots) |> 
          select(-Tcrit.mn, -T95.mn) # see summary output, only interested in T50

autoplot(htol_500_boots) # plots logistic curves and dashed lines for Tcrit, T50, T95 

#extracting the CI for the individual T50 found from bootstrapping
htol_500_boots_ci <- purrr::map_dfr(names(htol_500_boots$results), function(id) {
  t50_boots <- htol_500_boots$results[[id]]$T50
  data.frame(
    id    = id,
    T50.mn  = mean(t50_boots, na.rm = TRUE),
    T50.low = quantile(t50_boots, 0.025, na.rm = TRUE),
    T50.hi  = quantile(t50_boots, 0.975, na.rm = TRUE)
  )
})

# should probably add columns for age_class, month back in before exporting 
# maybe CI too? 
# Export the hot T50 data when ready
write.csv(htol_summary_500_boots,
  here("raw_data", "T50_hot_500_boots.csv"),
  row.names = FALSE
)

```

## COLD

### Raw data scatter plot

Making scatter plot for visualization. Points for each individual plant represented by colour and connected by lines. Faceted by month x age class.

```{r}
#| label: cold visualization plot  
#| message: false
   
# summary vizualisation plot for cold treatment faceted by month x age class 
ggplot(cold_subset, aes(x = as.numeric(as.character(temp_exp)), y = fv_fm_final, 
                       colour = factor(plant_nr))) +  
  
  geom_line(alpha = 0.4) +  # one line per plant
  geom_point(alpha = 0.6, size = 1.5) + # one point per measurement
  facet_grid(month ~ age_class) + # facet by month x age class

# add labels  
  labs( 
    title = "Fv/Fm response across freezing treatments from 5 C to -20 C",
    subtitle = "Each colour and line represent one individual plant.", 
    x = "Treatment temperature (C)",
    y = "Fv/Fm",
    colour = "Plant nr"
  ) +

#set theme  
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) 

# saving plot to output folder
ggsave(here("output", "cold_scatter_line_plot.png"), 
       width = 10, height = 6, dpi = 300) 
```

Problematic November low starting values: November_Building_4 and November_Building_3

Other November notes: Newly Burnt_3, Pioneer_7, Pioneer_5, Building_2, Mature_2 fvfm dips and goes back up.

January low starting values - many - but this seems like a general trend. So maybe not problematic, but meaningful biologically. Many very bumpy lines, especially Pioneer_4 which dies at -15 and comes back to life.

### Exploratory: FREEZING RESPONSE CURVES AND T50 - group level

```{r}
#| label: psiiht for cold
#| message: false

# stiching columns together 
united_columns_cold <- unite(
  cold_subset,
  col = "new_id", 
  month, age_class, #plant_nr, # now fitting nls for each age_class x month
  sep = "_",
  remove = TRUE
)

set.seed(8) # set seed before every psiiht  
# estimating cold tolerance curves using psiiht function
ctol <- psiiht(temperature = united_columns_cold$temp_exp,
               warming = FALSE,
               fvfm = united_columns_cold$fv_fm_final,  
               control_temp = 5, 
               id = united_columns_cold$new_id, 
               boots = 50) 

ctol_summary_all <- summary(ctol) |> 
       select(-Tcrit.mn, -T95.mn) # see summary output, only interested in T50

autoplot(ctol) # plots logistic curves and dashed lines for Tcrit, T50, T95 


# struggles to fit lines to the data when run in(does not reach the T-values)
# how to deal with this - will end up with very little data if removing rows where nls cant be fit
```

Model only works of run on group level. Even for group level estimates, it is missing T50s for January mature and newly burnt. Entails that these are the most problematic to model. Makes sense when comparing to raw data plots, does not drop low enough to calculate T50. Will try to run model on month x age class subsets to see if that works.

#### Exploratory: Exponantial model

An exponential model could potentially work better based on the shape of the curves for cold data. Does NOT work. Chunk removed. NOTE 21/4: Decided to not move forward with that.

#### Exploratory: Running on month x age_class groups separately to see which ones break

```{r}
#| label: separate runs for ageclass x month groups
#| message: false

# Filter for for month + age class + eventually in/out individuals to check where the problems come from.
cold_test <- cold_subset |> 
  filter(month == "November", age_class == "Newly Burnt")

# update unite as well
united_columns_cold_test <- unite(
  cold_test,
  col = "new_id",
  month, age_class, plant_nr,
  sep = "_",
  remove = TRUE
)

# run psiiht
set.seed(8)
ctol_test <- psiiht(temperature = united_columns_cold_test$temp_exp,
                   warming = FALSE,
                   fvfm = united_columns_cold_test$fv_fm_final,
                   control_temp = 5,
                   id = united_columns_cold_test$new_id,
                   boots = 50)

ctol_summary_test <- summary(ctol_nov) |>
  select(-Tcrit.mn, -T95.mn)

autoplot(ctol_nov_test)

```

Age_class (number of successful estimates/total individual plants)

November - Newly Burnt (8/8), Pioneer (7/8), Building (7/8) , Mature ((7/8)need to filter out plant_nr 3 before psiiht) works on individual level

January - Newly burnt (0/8) works on no individuals, Pioneer (3/8) only works if filtering for only plant_nr 6, 7, 8. Building (1/8) only works on plant 4, Mature (2/8) only works if filtering out plant_nr 4.

After running on subsets of data, it seems like this is the only solution for running cold tolerance on individual plant level.

Will run model separately for each group with 500 boots, filtering out specific individual that break the model if needed.

Important to note that results are uncertain, need to report how many that does converge and the prop_success of those that actually do.

Will have to join the result data frames together after.

### 500 BOOTS COLD PSIIHT FOR FINAL T50:

```{r}
#| label: 500 boots cold T50
#| message: false
# make united columns object for each month x age class group
# filter out specific plant_nr that stops the model from running remaining estimates

# November - Newly Burnt (8/8 estimates expected to work)
united_nov_nb <- cold_subset |>
  filter(month == "November", age_class == "Newly Burnt") |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

# November - Pioneer (7/8 estimates expected to work)
united_nov_pioneer <- cold_subset |>
  filter(month == "November", age_class == "Pioneer") |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

# November - Building (7/8 estimates expected to work)
united_nov_building <- cold_subset |>
  filter(month == "November", age_class == "Building") |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

# November - Mature (7/8 estimates expected to work, if filtering out plant 3)
united_nov_mature <- cold_subset |>
  filter(month == "November", age_class == "Mature", !plant_nr %in% c(3)) |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

# January - Newly Burnt (0/8) - SKIPPED, no individuals converge

# January - Pioneer (4/8 estimates expected to work, if filtering for only nr 4, 6, 7, 8)
united_jan_pioneer <- cold_subset |>
  filter(month == "January", age_class == "Pioneer", plant_nr %in% c(6, 7, 8)) |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

# January - Building (1/8 estimates expected to work, if filtering for only nr 4)
united_jan_building <- cold_subset |>
  filter(month == "January", age_class == "Building", plant_nr %in% c(4)) |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)

# January - Mature (2/8 estimates expected to work, filtering out plant 4)
united_jan_mature <- cold_subset |>
  filter(month == "January", age_class == "Mature", !plant_nr %in% c(4)) |>
  unite(col = "new_id", month, age_class, plant_nr, sep = "_", remove = TRUE)


# running model with 500 boots for each group
# setting seed before each block so they can be reproducible if run independently

set.seed(8)
# November - Newly Burnt (8/8)
ctol_nov_nb <- psiiht(temperature = united_nov_nb$temp_exp,
                      warming = FALSE,
                      fvfm = united_nov_nb$fv_fm_final,
                      control_temp = 5,
                      id = united_nov_nb$new_id,
                      boots = 500)
summary_nov_nb <- summary(ctol_nov_nb) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "November", age_class = "Newly Burnt")

set.seed(8)
# November - Pioneer (7/8)
ctol_nov_pioneer <- psiiht(temperature = united_nov_pioneer$temp_exp,
                            warming = FALSE,
                            fvfm = united_nov_pioneer$fv_fm_final,
                            control_temp = 5,
                            id = united_nov_pioneer$new_id,
                            boots = 500)
summary_nov_pioneer <- summary(ctol_nov_pioneer) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "November", age_class = "Pioneer")

set.seed(8)
# November - Building (7/8)
ctol_nov_building <- psiiht(temperature = united_nov_building$temp_exp,
                             warming = FALSE,
                             fvfm = united_nov_building$fv_fm_final,
                             control_temp = 5,
                             id = united_nov_building$new_id,
                             boots = 500)
summary_nov_building <- summary(ctol_nov_building) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "November", age_class = "Building")

set.seed(8)
# November - Mature (7/8)
ctol_nov_mature <- psiiht(temperature = united_nov_mature$temp_exp,
                           warming = FALSE,
                           fvfm = united_nov_mature$fv_fm_final,
                           control_temp = 5,
                           id = united_nov_mature$new_id,
                           boots = 500)
summary_nov_mature <- summary(ctol_nov_mature) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "November", age_class = "Mature")


# January - Newly Burnt (0/8) works on individual level, model breaks. 

set.seed(8)
# January - Pioneer (3/8) still unstable, try to rerun a few times if not working
ctol_jan_pioneer <- psiiht(temperature = united_jan_pioneer$temp_exp,
                            warming = FALSE,
                            fvfm = united_jan_pioneer$fv_fm_final,
                            control_temp = 5,
                            id = united_jan_pioneer$new_id,
                            boots = 500)
summary_jan_pioneer <- summary(ctol_jan_pioneer) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "January", age_class = "Pioneer")

set.seed(8)
# January - Building (1/8)
ctol_jan_building <- psiiht(temperature = united_jan_building$temp_exp,
                             warming = FALSE,
                             fvfm = united_jan_building$fv_fm_final,
                             control_temp = 5,
                             id = united_jan_building$new_id,
                             boots = 500)
summary_jan_building <- summary(ctol_jan_building) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "January", age_class = "Building")

set.seed(8)
# January - Mature (2/8)
ctol_jan_mature <- psiiht(temperature = united_jan_mature$temp_exp,
                           warming = FALSE,
                           fvfm = united_jan_mature$fv_fm_final,
                           control_temp = 5,
                           id = united_jan_mature$new_id,
                           boots = 500)
summary_jan_mature <- summary(ctol_jan_mature) |>
  select(-Tcrit.mn, -T95.mn) |>
  mutate(month = "January", age_class = "Mature")


# Merge all summaries (excluding January Newly Burnt - which worked for 0/8)
ctol_psiiht_summary <- bind_rows(
  summary_nov_nb,
  summary_nov_pioneer,
  summary_nov_building,
  summary_nov_mature,
  summary_jan_pioneer,
  summary_jan_building,
  summary_jan_mature
)
# adding in NA for individuals that could not run
# adding in remark column
```

#### T50 data export

```{r}
#| label: T50 data export 
#| message: false

# Create rows for excluded individuals
excluded_individuals <- tibble(
  id = c(
    # January - Newly Burnt (0/8, all excluded)
    "January_Newly Burnt_1", "January_Newly Burnt_2", "January_Newly Burnt_3",
    "January_Newly Burnt_4", "January_Newly Burnt_5", "January_Newly Burnt_6",
    "January_Newly Burnt_7", "January_Newly Burnt_8",
    # January - Pioneer (5/8 excluded)
    "January_Pioneer_1", "January_Pioneer_2", "January_Pioneer_3", "January_Pioneer_4", "January_Pioneer_5",
    # January - Building (7/8 excluded)
    "January_Building_1", "January_Building_2", "January_Building_3",
    "January_Building_5", "January_Building_6", "January_Building_7", "January_Building_8",
    # January - Mature (plant 4 excluded, but only 2/8 converged so remaining 6 returned NA directly)
    "January_Mature_4",
    # November - Mature (plant 3 excluded)
    "November_Mature_3"
  ),
  T50.mn = NA,
  prop_success = NA,
  remark = "; filtered out before psiiht run to not break function"
)

# merge with main summary and assign remarks
ctol_all_summary_500_boots <- bind_rows(
  ctol_psiiht_summary |> mutate(remark = NA_character_),  # initialise remark column
  excluded_individuals
) |>
  arrange(id) |>
  mutate(remark = case_when(
    is.nan(prop_success) ~ "psiiht convergence failure",
    is.na(T50.mn) & is.na(prop_success) ~ "filtered out before psiiht run to not break function",
    .default = remark
  ))|>
  select(-month, -age_class)

# export the cold T50 data to raw_data folder
write.csv(ctol_all_summary_500_boots,
  here("raw_data", "T50_cold_500_boots.csv"),
  row.names = FALSE
)
```

## **T50 DATA ANALYSIS**

### HOT

```{r}
#| label: T50 data import
#| message: false

T50_hot <- read.csv(here("raw_data", "T50_hot_500_boots.csv"))

```

```{r}
#| label: T50 hot data checking 
#| message: false

summary(T50_hot)              # global check of all individuals
colSums(is.na(T50_hot))      # see if a lot of NA, there are two NAs


# No values outside the expected range. Two individuals cannot return T50, is NA.

# RICHARD SAID TO ONLY REMOVE IMPOSSIBLE VALUES, SOME VARIATION IS ALWAYS EXPECTED 
# RICHARD SAID NO NEED TO CHECK FOR NORMALITY AS DATASET IS SMALL, WILL PROBABLY NOT BE NORMAL
# JUST MOVE ON TO LINEAR MODELS

# visualising prop_succsess vs etimated T50
T50_hot |>
  filter(!is.na(T50.mn)) |>
  ggplot(aes(x = prop_success, y = T50.mn)) +
  geom_point() +
  labs(title = "T50 vs bootstrap success rate")

# testing for T50 outliers using z scores, none detected
T50_hot |>
  filter(!is.na(T50)) |>
  mutate(z_score = scale(T50)) |>
  filter(abs(z_score) > 3)
```

All T50 estimates are in the rage between 49.55 C - 55.27 C, this is reasonable. Two individuals returns NA, this is where the model could not fit T50 estimates.

#### Preparing hot T50 data for analysis

```{r}
#| label: preparing T50 data 
#| message: false
#adding columns for month, age class, id for analysis
T50_hot <- T50_hot |> 
  separate(
    id,
    into = c("month", "age_class", "plant_nr"),
    sep = "_",
    remove = FALSE   # keeps the original id column
  ) |> 
  mutate(plant_nr = as.numeric(plant_nr)) |> # plant number as numeric
  rename(T50 = T50.mn)|> # rename T50 column
  na.omit() # remove rows with NA

```

#### Hot group level visualization plot

```{r}
#| label: group level visualisation 
#| message: false

# grouping T50 by month x age_class
T50_hot_grouped <- T50_hot |> 
  group_by(month, age_class) |> 
  mutate(month = factor(month, levels = c("November", "January"))) |> 
  mutate(age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer", 
                                  "Building", "Mature"))) |> 
    summarise(
    n = n(), # see how many individuals are used in calculating mean T50
    mean_T50 = mean(T50),
    SD = sd(T50),
    SE = SD / sqrt(n),
    .groups = "drop"
  )
# n = 7 for January x pioneer and January x building, all others have n = 8 

# visualisation plot to see how T50 varies with month and age class and possible interactions

ggplot(T50_hot_grouped, aes(x = month, y = mean_T50, color = age_class, group = age_class)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = mean_T50 - SE, ymax = mean_T50 + SE), width = 0.2) +
  theme_bw()

```

#### Statistics - Hot T50 linear model

Using linear models. T50_hot and T50_cold as response variable for each.

Fixed effects: month and age_class. Random effect: none as each plant is reduced to one T50-value now, the random effect is not needed.

Using a two-way ANOVA

Using a model with interaction to take hypothesis 3 into account. No interaction found, comparing interaction model with no-interaction model.

```{r}
#| label: hot linear model 
#| message: false

# linear model for T50 with month and age class as fixed effects. 
# interaction needed to test hypothesis 3

hot_model_interaction <- lm(T50 ~ month * age_class, data = T50_hot) # interaction 

anova(hot_model_interaction) #look at and report results
summary(hot_model_interaction) # look at and report results


# month highly significant (p < 0.001). Hypothesis 1 supported. 
# age class is not significant (p = 0.18) 
# month:age_class interaction is non-significant (p = 0.68). Hypothesis 3 not supported. 

# since interaction is not significant, trying model without to see if it is better
hot_model <- lm(T50 ~ month + age_class, data = T50_hot) # no interaction

anova(hot_model) #look at and report results
summary(hot_model) # look at and report results

# summed up: heat tolerance appox 2 C higher in January than in November. 
# month have a significant effect on T50 hot***
# age class have NO significant effect on T50 hot

#comparing model fits 
anova(hot_model, hot_model_interaction)
# model not significantly removed by adding interaction

AIC(hot_model, hot_model_interaction) 
# AIC 195 for nointeraction, 199 for interaction
```

By both models: Hypothesis 1 is supported, month have significant effect on T50. No significant effect of age class on T50.

Model with interaction: No interaction between month and age_class, hypothesis 3 NOT supported.

#### Checking model performance

```{r}
#| label: checking model preformance 
#| message: false

library(performance)
check_model(hot_model_interaction)  
# collionearity is a clear problem here. 
# the interaction term increase variance in parameter estimates, making coefficients less reliable. 
# better to use the model with no interaction!


check_model(hot_model)  
# all panels good; except some slight deviations form whats excpected in influential observations. 
# still within reasonable limits. 

```

Results from performance model check is OK for model WITHOUT interaction. I would opt for that one as it is simpler. Need confirmation if this is ok, or if the interaction model is fine anyway.

#### Export model result in a table

```{r}

```

### COLD

```{r}
#| label: T50 cold data import
#| message: false
# importing full dataframe for cold results run with 500 boots

T50_cold <- read.csv(here("raw_data", "T50_cold_500_boots.csv"))

```

```{r}
#| label: T50 hot data checking 
#| message: false

summary(T50_cold)              # global check of all individuals, see if any are very high or low

# big variation in estimated ctol, from -41.23 to -13.84 C. 
# some individuals have very low prop_success, mening model failed to converge, giving less reliable estimates

# visualising prop_succsess vs etimated T50 
T50_cold |>
  filter(!is.na(T50.mn)) |>
  ggplot(aes(x = prop_success, y = T50.mn)) +
  geom_point() +
  geom_hline(yintercept = -35, linetype = "dashed", colour = "red") +  # adjust threshold
  labs(title = "T50 vs bootstrap success rate")

# testing for T50 outliers using z-scores. one detected, november building 6. 
T50_cold |>
  filter(!is.na(T50.mn)) |>
  mutate(z_score = scale(T50.mn)) |>
  filter(abs(z_score) > 3) 
```

Some individuals have very low estimated T50 (\<-35 C) and also low prop_success. Would be inclined to filter out, as these results are unreliable. Colud affect linear model results substantially.

Z-score over 3 SD for November_Building_6 (T50 = 41C). Would also consider filtering out January_Mature_8 as tolerance deviates almost just as much with T50 = 39 C.

No individuals have high prop_success though so would have very few individuals left if being strict on this.

Need to come back and ask if this is justified.

#### Preparing cold T50 data for analysis and outlier handling

```{r}
#| label: preparing T50 cold data
#| message: false
#adding columns for month, age class, id for analysis

T50_cold <- T50_cold |>
  separate(
    id,
    into = c("month", "age_class", "plant_nr"),
    sep = "_",
    remove = FALSE
  ) |>
  mutate(plant_nr = as.numeric(plant_nr)) |>
  rename(T50 = T50.mn) |>
  # add remarks before filtering
  mutate(remark = case_when(
    T50 < -35 ~ "; outlier low prop_succes and very low T50 predicted",
    .default = remark
  )) |>
  # now filter out
  filter(T50 >= -35) |> # filter out as low prop_success, and very low t50
  filter(!is.na(T50))  # only remove rows where T50 itself is NA
  
```

#### Cold group level visualization plot

```{r}
#| label: group level visualisation 
#| message: false

# grouping T50 by month x age_class
T50_cold_grouped <- T50_cold |> 
  group_by(month, age_class) |> 
  mutate(month = factor(month, levels = c("November", "January"))) |> 
  mutate(age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer", 
                                  "Building", "Mature"))) |> 
    summarise(
    n = n(), # see how many individuals are used in calculating mean T50
    mean_T50 = mean(T50),
    SD = sd(T50),
    SE = SD / sqrt(n),
    .groups = "drop"
  )
# n = 8 only for one group. November plants have n in rage from 6-8. January n between 0 and 3.  

# visualisation plot to see how T50 varies with month and age class and possible interactions

ggplot(T50_cold_grouped, aes(x = month, y = mean_T50, color = age_class, group = age_class)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = mean_T50 - SE, ymax = mean_T50 + SE), width = 0.2) +
  theme_bw()
```

#### Statistics - Hot T50 linear model

Not sure how to prceed as there are very few estimates for January T50s. Only 5 individuals accross all of January got a T50. Only January_Pioneer got more than one individual T50, so only one with SD and SE estimation possible.

Any statistical comparison between months would be very fragile.

Any point of fitting a model across months here with these uncertainties? Alternatively just report January result in table and try to run a model for November alone? Would then not be able to test H1 at all, and only partially test H3.

```{r}
#| label: cold linear mixed effect model 
#| message: false

# if runn on all available cold T50 data, with interaction
cold_model <- lm(T50 ~ month * age_class, data = T50_cold)
anova(cold_model)
summary(cold_model)
# not a good model, too much uncertainty in Jnauary. Newly Burnt T50 missing entirely. 


# could run model on ONLY November T50s, then not testing H1 at all, and H3 only partially
T50_cold_nov <- T50_cold |>
  filter(month == "November")

# linear model for cold T50 with age class as fixed effect
cold_model_nov <- lm(T50 ~ age_class, data = T50_cold_nov)
anova(cold_model_nov)
summary(cold_model_nov)


# age class does not significantly affect cold tolerance in November (p = 0.29, adjusted R = 0.03).
```

Still useful to connect to traits? If so only for November?

#### Export model result in a table?

------------

#### Plot response curves and T50 for EGU and thesis

```{r}
#| label: hot preparing curves and T50 mean
#| message: false

# currently for the test chunk htol 20 all

# EXTRACT PREDICTED CURVES FROM RESULTS OBJECT 
# each individual in htol_500_boots$results has a $preds dataframe with
# columns: temperature, preds (fitted curve), low, high (bootstrap CI band)

pred_curves <- purrr::map_dfr(names(htol_500_boots$results), function(id) {
  preds <- htol_500_boots$results[[id]]$preds
  if (is.null(preds)) return(NULL)
  preds |>
    mutate(id = id)
}) |>
  # Split id back into month, age_class, plant_nr
  separate(id, into = c("month", "age_class", "plant_nr"), sep = "_")


# AVERAGE CURVES TO GROUP LEVEL FOR PLOTTING 

group_curves <- pred_curves |>
  mutate(
    month     = factor(month, levels = c("November", "January")),
    age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer",
                                              "Building",   "Mature"))
  ) |>
  group_by(month, age_class, temperature) |>
  summarise(
    mean_pred = mean(preds, na.rm = TRUE),
    .groups   = "drop"
  )

# PREPARE GROUP-LEVEL T50 WITH CI 

t50_points <- purrr::map_dfr(names(htol_500_boots$results), function(id) {
  t50_boots <- htol_500_boots$results[[id]]$T50
  if (is.null(t50_boots)) return(NULL)
  data.frame(
    id       = id,
    T50.mn   = mean(t50_boots,                      na.rm = TRUE),
    T50.low  = quantile(t50_boots, 0.025,            na.rm = TRUE),
    T50.hi   = quantile(t50_boots, 0.975,            na.rm = TRUE)
  )
}) |>
  separate(id, into = c("month", "age_class", "plant_nr"), sep = "_") |>
  mutate(
    month     = factor(month,     levels = c("November", "January")),
    age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer",
                                              "Building",   "Mature"))
  ) |>
  # Summarise to group level
  group_by(month, age_class) |>
  summarise(
    mean_T50 = mean(T50.mn,  na.rm = TRUE),
    low_T50  = mean(T50.low, na.rm = TRUE),
    hi_T50   = mean(T50.hi,  na.rm = TRUE),
    .groups  = "drop"
  ) |>
  # Find Fv/Fm value on the curve at the T50 temperature
  left_join(group_curves, by = c("month", "age_class")) |>
  group_by(month, age_class, mean_T50, low_T50, hi_T50) |>
  slice_min(abs(temperature - mean_T50)) |>
  ungroup()



```

```{r}
#| label: hot plot curve t50 faceted by age class
#| message: false
# PLOT

ggplot(group_curves, aes(x = temperature, y = mean_pred,
                          linetype = month, colour = month)) +

  geom_line(linewidth = 0.9) +

  # solid November, dashed January
  scale_linetype_manual(values = c("November" = "solid",
                                   "January"  = "solid")) +
  scale_colour_manual(values = c("November" = "#7B2D8B",   # dark purple
                                "January"  = "#C589D8")) + # light purple

    # Horizontal error bar showing 95% CI along x-axis
  geom_errorbarh(data = t50_points,
                 aes(y      = mean_pred,
                     xmin   = low_T50,
                     xmax   = hi_T50,
                     colour = month),
                 height = 0.02, linewidth = 0.6,
                 show.legend = FALSE) +

  # T50 point — now mapped to shape so it appears in legend
  geom_point(data = t50_points,
             aes(x = mean_T50, y = mean_pred,
                 colour = month, shape = "T50"),
             size = 3.5, show.legend = TRUE) +
  

  # Dotted vertical drop line to x-axis
  geom_segment(data = t50_points,
               aes(x      = mean_T50, xend = mean_T50,
                   y      = 0,        yend = mean_pred,
                   colour = month),
               linetype = "dotted", linewidth = 0.4,
               show.legend = FALSE) +
  
  # label on T50 pionts
  geom_text(data = t50_points,
            aes(x      = mean_T50,
                y      = mean_pred,
             label = paste0(ifelse(month == "November", "Nov", "Jan"),
               ", ", round(mean_T50, 1), "C"),
                colour = month),
            vjust       = ifelse(t50_points$month == "November", 2, -2),
            hjust       = ifelse(t50_points$month == "November", 1.2, 0.1),
            size        = 2.5,
            fontface    = "bold",
            show.legend = FALSE) +

  # 2x2 facet grid in correct positions:
  # top-left = Newly Burnt, top-right = Pioneer
  # bottom-left = Building, bottom-right = Mature
  facet_wrap(~ age_class,
             ncol = 2,
             nrow = 2) +

  labs(
    x        = "Temperature (C)",
    y        = "Fv/Fm",
    linetype = NULL,
    colour   = NULL,
    shape    = NULL
  ) +

  scale_y_continuous(limits = c(0, 0.9)) +
  
   # Tidy up the legend so lines and point appear cleanly
  guides(
    colour   = guide_legend(override.aes = list(shape = NA)),
    shape    = guide_legend(override.aes = list(colour = "grey30",
                                                linetype = "blank"))
  ) +
  
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )


# save fig to output folder 
ggsave(
  here("output", "T50_hot_group_curves_age_class.png"),
  width  = 10,
  height = 6,
  dpi    = 300
)
```

```{r}
#| label: hot plot curve t50 faceted by age class
#| message: false
# PLOT

ggplot(group_curves, aes(x = temperature, y = mean_pred,
                          linetype = month, colour = month)) +

     # Raw data points — small and transparent
  geom_point(data = hot_subset |>
               mutate(month     = factor(month, levels = c("November", "January")),
                      age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer",
                                                                "Building", "Mature"))),
             aes(x = temp_exp, y = fv_fm_final, colour = month),
             size = 0.8, alpha = 0.3, show.legend = FALSE) +

  geom_line(linewidth = 0.9) +

  # solid November, dashed January
  scale_linetype_manual(values = c("November" = "solid",
                                   "January"  = "solid")) +
  scale_colour_manual(values = c("November" = "#7B2D8B",   # dark purple
                                "January"  = "#C589D8")) + # light purple

    # Horizontal error bar showing 95% CI along x-axis
  geom_errorbarh(data = t50_points,
                 aes(y      = mean_pred,
                     xmin   = low_T50,
                     xmax   = hi_T50,
                     colour = month),
                 height = 0.02, linewidth = 0.6,
                 show.legend = FALSE) +

  # T50 point — now mapped to shape so it appears in legend
  geom_point(data = t50_points,
             aes(x = mean_T50, y = mean_pred,
                 colour = month, shape = "T50"),
             size = 3.5, show.legend = TRUE) +
  

  # Dotted vertical drop line to x-axis
  geom_segment(data = t50_points,
               aes(x      = mean_T50, xend = mean_T50,
                   y      = 0,        yend = mean_pred,
                   colour = month),
               linetype = "dotted", linewidth = 0.4,
               show.legend = FALSE) +
  
  # Horizontal dotted line at T50 Fv/Fm value for each group
  geom_hline(data = t50_points,
             aes(yintercept = mean_pred, colour = month),
             linetype = "dotted", linewidth = 0.4,
             show.legend = FALSE) +
  
  # label on T50 pionts
  geom_text(data = t50_points,
            aes(x      = mean_T50,
                y      = mean_pred,
             label = paste0(ifelse(month == "November", "Nov", "Jan"),
               ", ", round(mean_T50, 1), "C"),
                colour = month),
            vjust       = ifelse(t50_points$month == "November", 2, -2),
            hjust       = ifelse(t50_points$month == "November", 1.2, 0.1),
            size        = 2.5,
            fontface    = "bold",
            show.legend = FALSE) +

  # 2x2 facet grid in correct positions:
  # top-left = Newly Burnt, top-right = Pioneer
  # bottom-left = Building, bottom-right = Mature
  facet_wrap(~ age_class,
             ncol = 2,
             nrow = 2) +

  labs(
    x        = "Temperature (C)",
    y        = "Fv/Fm",
    linetype = NULL,
    colour   = NULL,
    shape    = NULL
  ) +

  scale_y_continuous(limits = c(0, 0.9)) +
  
   # Tidy up the legend so lines and point appear cleanly
  guides(
    colour   = guide_legend(override.aes = list(shape = NA)),
    shape    = guide_legend(override.aes = list(colour = "grey30",
                                                linetype = "blank"))
  ) +
  
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )


# save fig to output folder 
ggsave(
  here("output", "T50_hot_group_curves_age_class_test.png"),
  width  = 8,
  height = 6,
  dpi    = 300
)
```

```{r}

```
