---
title: "Traits"
format: html
editor: visual
---

## Loading library

```{r}
#| label: loading library
#| message: false

library(tidyverse)
library(here)
library(ggplot2)
library(reshape2)

```

## Data import

```{r}
#| label: data import 
#| message: false

# clean data import, selecting proj_2025 = traits, and only relevant columns
trait_data <- read.csv(here("clean_data", "DURIN_Clean_Seasonal_Thermal_Tolerance_Bente_Msc.csv")) |>
  dplyr::filter(proj_2025 == "Traits") |>
  dplyr::select(proj_2025, month, age_class, plant_nr, 
                                remark, leaf_thickness_mean, sla, ldmc) |> 
  dplyr::mutate(id = paste(month, age_class, plant_nr, sep = "_"))

# implort hot T50 data
T50_hot_data <- read.csv(here("raw_data", "T50_hot_500_boots.csv")) |> 
   rename(T50 = T50.mn)  # rename T50 column


# implort hot T50 data
T50_cold_data <- read.csv(here("raw_data", "T50_cold_500_boots.csv")) |> 
   rename(T50 = T50.mn)  # rename T50 column

```

## Mean table and exploratory plot traits

More for practice of the workflow. Was testing my alternative hypothesis nr 2, but I do not think it is useful for more than supplementary material in any case. Plot might be useful still, so keeping it for now.

```{r}
#| label: exploratory plot 
#| message: false

# make trait object
traits <- c("leaf_thickness_mean", "sla", "ldmc")
 
# tidy labels for facet strips
trait_labels <- c(
  leaf_thickness_mean = "Leaf Thickness (mm)",
  sla                 = "SLA (cm/g)",
  ldmc                = "LDMC (mg/g)"
)
 
df_traits <- trait_data |>
  filter(proj_2025 == "Traits") |>
  select(plant_nr, age_class, month, all_of(traits)) |>   # keep relevant cols
  mutate(
    month     = factor(month,     levels = c("November", "January"), # adjusting levels for fig
                       labels = c("November", "January")),
    age_class = factor(age_class, levels = c( "Newly Burnt", "Pioneer", 
                                             "Building", "Mature"))                           
  )


# want a summary table of means of trait values  pr group
summary_table <- df_traits |>
  group_by(age_class, month) |>
  summarise(
    thickness_mean = round(mean(leaf_thickness_mean, na.rm = TRUE), 4),
    thickness_sd   = round(sd(leaf_thickness_mean,   na.rm = TRUE), 4),
    sla_mean       = round(mean(sla,                 na.rm = TRUE), 4),
    sla_sd         = round(sd(sla,                   na.rm = TRUE), 4),
    ldmc_mean      = round(mean(ldmc,                na.rm = TRUE), 4),
    ldmc_sd        = round(sd(ldmc,                  na.rm = TRUE), 4),
    n              = n(),
    .groups = "drop"
  )

# View it
print(summary_table)

# Export as CSV
#write.csv(summary_table, "trait_summary.csv", row.names = FALSE)
 
# pivot to long format
df_long <- df_traits |>
  pivot_longer(cols = all_of(traits),
               names_to  = "trait",
               values_to = "value") |>
  mutate(trait = factor(trait, levels = traits))
 
# making plot 
traits_plotted <- ggplot(df_long,
            aes(x = age_class, y = value, fill = month, colour = month)) +
 
  # raw points (jittered, dodged to match boxes)
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    size  = 1.6, alpha = 0.45, shape = 16
  ) +
 
  # box plots (dodged, slightly transparent)
  geom_boxplot(
    position     = position_dodge(width = 0.75),
    width        = 0.55,
    outlier.shape = NA,          # hide outlier dots (already shown as points)
    alpha        = 0.35,
    linewidth    = 0.55
  ) +
 
  # mean ± 1 SD error bars
  stat_summary(
    fun.data = mean_se, # standard error around mean
    geom      = "errorbar",
    position  = position_dodge(width = 0.75),
    width     = 0.25,
    linewidth = 0.7
  ) +
 
  # mean dot 
  stat_summary(
    fun      = mean, 
    geom     = "point",
    shape    = 23,               # filled diamond
    size     = 2.5,
    position = position_dodge(width = 0.75)
  ) +
 
  # facet: one panel per trait 
  facet_wrap(~ trait,
             scales   = "free_y",
             labeller = as_labeller(trait_labels),
             ncol     = 3) +
 
  # colour
  scale_fill_manual(  values = c("November" = "#E07B4F", "January" = "#4E9EC2")) +
  scale_colour_manual(values = c("November" = "#E07B4F", "January" = "#4E9EC2")) +
 
  # labels
  labs(
    title  = "Leaf Trait Measurements by Age Class and Month",
    x      = "Age Class",
    y      = "Trait Value",
    fill   = "Month",
    colour = "Month"
  ) +
 
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "grey92", colour = "grey60"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "top",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 30, hjust = 1)
  )

print(traits_plotted) # must be run in the console

# add ggsave when happy
 
```

## Trait means for each plant individual

Before combining with T50 data, have to reduce the 5 measurements per individual to one mean value. As T50 data only have one value per individual.

```{r}
#| label: means per individual plant  
#| message: false

# collapsing the five (or four) measurements of traits per individual id plant, into the mean of them
trait_data_means <- trait_data |>
  dplyr::group_by(id, month, age_class, plant_nr) |>
  dplyr::summarise(
    leaf_thickness_mean = mean(leaf_thickness_mean, na.rm = TRUE),
    sla = mean(sla, na.rm = TRUE),
    ldmc = mean(ldmc, na.rm = TRUE),
    .groups = "drop"
  )

```

## Correlation matrix between traits

Only problematic if correlation is above 0.7-0.9.

```{r}
#| label: correlation matrix   
#| message: false

# correlation matrix NOT taking month into account 
trait_cor <- trait_data_means |>
  dplyr::select(leaf_thickness_mean, sla, ldmc) |>
  cor(use = "complete.obs")

trait_cor # all within +/- 0.7 so no problematic correlations for lm model later

#correlation matrix taking month into account
trait_cor_by_month <- trait_data_means |>
  dplyr::group_by(month) |>
  dplyr::group_map(~ {
    cor(
      dplyr::select(.x, leaf_thickness_mean, sla, ldmc),
      use = "complete.obs"
    )
  })

trait_cor_by_month #  all within +/- 0.7 so no problematic correlations for lm model later
# sla and ldmc moderately negatively correlated around -0.6, as expected 

```

## HOT

## Joining T50 hot data to trait data

```{r}
#| label: joining data frames   
#| message: false

# joining tarit means data frame with T50 hot dataframe by id column
trait_T50_hot <- trait_data_means |>
  dplyr::left_join(T50_hot_data, by = "id") |> 
   mutate(month = factor(month, levels = c("November", "January"))) |> 
  na.omit() # remove rows with NA

```

## Exploratory plots: T50 hot against each trait

```{r}
#| label: exploratory plots    
#| message: false

# tried without facteting by month. already know month has an impact, so could mask trait trends 
# faceting by month, as that cold mask trends 

# mean thickness vs T50 hot
thickness_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = leaf_thickness_mean, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "Leaf thickness", y = "T50") +
  theme_minimal()

thickness_vs_T50_hot 
# when not faceted by month: slight increase in T50 hot with increased thickness
# as expected
# when faceted by month: January relative flat relationship, November increased thickness corresponds to decreased T50. not as expected. 

# mean sla vs T50 hot
sla_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = sla, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "SLA", y = "T50") +
  theme_minimal()

sla_vs_T50_hot 
# when NOT faceted by month: slight increase in T50 hot with increased sla,
# not as expected. maybe the very high sla group jan_nb_1 influences this
# when faceted by month: January increased sla corresponds to increased T50, not expected. November flat relationship


# mean ldmc vs T50 hot
ldmc_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = ldmc, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "LDMC", y = "T50") +
  theme_minimal()

ldmc_vs_T50_hot 
# when NOT faceted by month: decrease in T50 hot with increased ldmc, not expected
# when faceted by month: Janury increased ldmc relate to decreased T50, not expected. November increased ldmc relate to slight increase in T50. 

```

## Linear model for traits vs T50 hot

Checking if traits correlate with T50 **independent** of time and age class.

```{r}
#| label: linear model   
#| message: false

# linear model for T50 response to triats, no interactions included
lm_traits_T50_hot <- lm(T50 ~ sla + leaf_thickness_mean + ldmc + month + age_class, data = trait_T50_hot) 

anova(lm_traits_T50_hot) #look at and report results
summary(lm_traits_T50_hot) # look at and report results

# summed up: no significant effects of any of these traits, again no effect of age class.
# as for original T50 model: month significant predictor!
```

## Checking model performance

```{r}
#| label: checking model preformance 
#| message: false

library(performance)
check_model(lm_traits_T50_hot)  

# all panels look good
# especially important to note: was concerned about month possibly masking effects of traits, but based on relatively low collinearity this should not be a problem
# model might hold as is
```

NOTE/QUESTION: Despite the model diagnostics looking good, the plots of traits vs T50 faceted by month show relationships in opposite directions. Is this reason to worry? Need to investigate.

## Model comparisons - with/without interactions

```{r}
#| label: comparing models
#| message: false

# checking if adding interctions makes the model better
# month interaction taken into account for all traits
lm_T50_interactions <- lm(
  T50 ~ (sla + leaf_thickness_mean + ldmc) * month + age_class,
  data = trait_T50_hot
)
summary(lm_T50_interactions)

model_main <- lm(
  T50 ~ sla + leaf_thickness_mean + ldmc + month + age_class,
  data = trait_T50_hot
)

anova(model_main, lm_T50_interactions) # p-value above 0.05, not significant. Not useful to take interactions for all traits into account. 

# only interaction for thickness * month
lm_thick_int <- lm(
  T50 ~ leaf_thickness_mean * month + sla + ldmc + age_class,
  data = trait_T50_hot
)

anova(model_main, lm_thick_int) # p-value above 0.05, not significant. Not useful to take interactions for all traits into account. 

# only interaction for sla * month
lm_sla_int <- lm(
  T50 ~ sla * month + leaf_thickness_mean + ldmc + age_class,
  data = trait_T50_hot
)

anova(model_main, lm_sla_int) # p-value above 0.05, not significant. Not useful to take interactions for all traits into account. 

# only interaction for ldmc * month
lm_ldmc_int <- lm(
  T50 ~ ldmc * month + sla + leaf_thickness_mean + age_class,
  data = trait_T50_hot
)

anova(model_main, lm_ldmc_int) # p-value under 0.05, significant, barely. Adding an interaction term here makes the model better.
# should consider including the interaction in the main model. 

check_model(lm_ldmc_int) # high collinearity, may inflate uncertanty. 
#preformance better if model without interaction

AIC(lm_traits_T50_hot, lm_ldmc_int) # slightly better for the interaction included
```

NEED TO DECIDE IF I WANT TO ADD INTERACTION INTO MODEL. Come back to this.

## Export model output

```{r}

```

## Clean final visualization plot

```{r}

# plot for each trait x t50 hot. facet by month and trait, giving 6 panels. 
```

HUSK: LOOKING AT T50 RELATED TO TRAITS - men uavhengig av måned og aldersklasse! Så må basere modellen på det.

## COLD

No/very unstable data for January individuals! possible to do linear model triats analysis still? if so, only on November

## Joining T50 cold data to trait data

```{r}
#| label: joining data frames   
#| message: false

# joining tarit means data frame with T50 hot dataframe by id column
trait_T50_cold <- trait_data_means |>
  dplyr::left_join(T50_cold_data, by = "id") |> 
   mutate(month = factor(month, levels = c("November", "January"))) |> 
   # now filter out extremely low, low_prop T50 estimates
   filter(T50 >= -35) |> 
   filter(!is.na(T50))  # only remove rows where T50 itself is NA

```

## Exploratory plots: T50 cold against each trait

```{r}
#| label: exploratory plots    
#| message: false

# faceting by month, as data availability 

# mean thickness vs T50 cold
thickness_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = leaf_thickness_mean, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "Leaf thickness", y = "T50") +
  theme_minimal()

thickness_vs_T50_cold 
# when not faceted by month: thicker leaves less tolerant to cold. 
# when faceted by month: November, slight increase in thickness correlated to less cold tolerant. January, clearer similar trend NB: very few data points. 

# mean sla vs T50 cold
sla_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = sla, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "SLA", y = "T50") +
  theme_minimal()

sla_vs_T50_cold 
# when NOT faceted by month: higher SLA (more aqusitive) have less cold tolerant leaves. as expected
# when faceted by month: November higher SLA (more aqusitive) have less cold tolerant leaves. as expected. January, opposite pattern, high SLA (aquisitive) correlates to better cold tolerance - NB few data points 

# mean ldmc vs T50 cold
ldmc_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = ldmc, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "LDMC", y = "T50") +
  theme_minimal()

ldmc_vs_T50_cold
# when NOT faceted by month: high LDMC correlates to slightly less cold tolerant leaves. not expected
# when faceted by month: November high LDMC are more cold tolerant. Janyary almost flat relationship - NB few data points. 

```

## Linear model for traits vs T50 cold

```{r}
#| label: linear model cold all
#| message: false

# linear model for T50 response to triats, no interactions included
lm_traits_T50_cold <- lm(T50 ~ sla + leaf_thickness_mean + ldmc + month + age_class, data = trait_T50_cold) 

anova(lm_traits_T50_cold) #look at and report results
summary(lm_traits_T50_cold) # look at and report results

# summed up: 
# model significant
#leaf thickness significant prdictor of T50. month effect likely underrepresented as very few January individuals are included. 
```

```{r}
#| label: linear model cold nov
#| message: false
lm_traits_T50_cold_nov <- lm(T50 ~ sla + leaf_thickness_mean + ldmc + age_class,
                            data = trait_T50_cold |> filter(month == "November"))
anova(lm_traits_T50_cold_nov)
summary(lm_traits_T50_cold_nov)

# summed up: 
# model lost significance whe january removed 
# leaf thickness not significant anymore
# probably bc of unbalanced data. 
```

## Checking model performance

```{r}
#| label: checking model preformance 
#| message: false

library(performance)
check_model(lm_traits_T50_cold)  # linearity, homogenity not very good

check_model(lm_traits_T50_cold_nov)  # a bit better 
```

Already low n = 28 and 5/6 predictors, so maybe not wise to add interactions.

```{r}
#| label: comparing models
#| message: false

# month x ldmc (mirrors hot model)
lm_cold_ldmc_int <- lm(T50 ~ ldmc * month + sla + leaf_thickness_mean + age_class,
                        data = trait_T50_cold)

# month x leaf thickness (biologically motivated given leaf thickness was borderline)
lm_cold_lt_int <- lm(T50 ~ leaf_thickness_mean * month + sla + ldmc + age_class,
                      data = trait_T50_cold)

# month x sla
lm_cold_sla_int <- lm(T50 ~ sla * month + leaf_thickness_mean + ldmc + age_class,
                       data = trait_T50_cold)

# compare all
AIC(lm_traits_T50_cold,
    lm_cold_ldmc_int,
    lm_cold_lt_int,
    lm_cold_sla_int)

# none of these options are better

```
