---
title: "Traits"
format: html
editor: visual
---

## Loading library

```{r}
#| label: loading library
#| message: false

library(tidyverse)
library(here)
library(ggplot2)
library(reshape2)
library(broom)
```

## Data import

Only include dry weight and hieght if time.

```{r}
#| label: data import 
#| message: false

# clean data import, selecting proj_2025 = traits, and only relevant columns
trait_data <- read.csv(here("clean_data", "DURIN_Clean_Seasonal_Thermal_Tolerance_Bente_Msc.csv")) |>
  dplyr::filter(proj_2025 == "Traits") |>
  dplyr::select(proj_2025, month, age_class, plant_nr, 
                                remark, leaf_thickness_mean, dry_mass_g, height_s_top, sla, ldmc) |> 
  dplyr::mutate(id = paste(month, age_class, plant_nr, sep = "_")) |> 
  dplyr::filter(!str_detect(remark, regex("outlier", ignore_case = TRUE))) # filter out flagged outliers

# import hot T50 data
T50_hot_data <- read.csv(here("raw_data", "T50_hot_500_boots.csv")) |> 
   rename(T50 = T50.mn)  # rename T50 column


# import hot T50 data
T50_cold_data <- read.csv(here("raw_data", "T50_cold_500_boots.csv")) |> 
   rename(T50 = T50.mn)  # rename T50 column

```

## Mean table and exploratory plot traits

More for practice of the workflow. Was testing my alternative hypothesis nr 2, but I do not think it is useful for more than supplementary material in any case. Plot might be useful still, so keeping it for now.

```{r}
#| label: exploratory plot 
#| message: false

# make trait object
traits <- c("dry_mass_g", "leaf_thickness_mean", "sla", "ldmc", "height_s_top")
 
# tidy labels for facet strips
trait_labels <- c(
  dry_mass_g          = "Leaf Dry Mass (g)",
  leaf_thickness_mean = "Leaf Thickness (mm)",
  sla                 = "SLA (cm/g)",
  ldmc                = "LDMC (mg/g)",
  height_s_top        = "Plant Height (cm)"
)
 
df_traits <- trait_data |>
  filter(proj_2025 == "Traits") |>
  select(plant_nr, age_class, month, all_of(traits)) |>   # keep relevant cols
  mutate(
    month     = factor(month,     levels = c("November", "January"), # adjusting levels for fig
                       labels = c("November", "January")),
    age_class = factor(age_class, levels = c( "Newly Burnt", "Pioneer", 
                                             "Building", "Mature"))                           
  )


# want a summary table of means of trait values  pr group
summary_table <- df_traits |>
  group_by(age_class, month) |>
  summarise(
    drymass_mean   = round(mean(dry_mass_g, na.rm = TRUE), 4),
    drymass_sd     = round(sd(dry_mass_g,   na.rm = TRUE), 4),
    thickness_mean = round(mean(leaf_thickness_mean, na.rm = TRUE), 4),
    thickness_sd   = round(sd(leaf_thickness_mean,   na.rm = TRUE), 4),
    sla_mean       = round(mean(sla,                 na.rm = TRUE), 4),
    sla_sd         = round(sd(sla,                   na.rm = TRUE), 4),
    ldmc_mean      = round(mean(ldmc,                na.rm = TRUE), 4),
    ldmc_sd        = round(sd(ldmc,                  na.rm = TRUE), 4),
    height_mean    = round(mean(height_s_top, na.rm = TRUE), 4),
    height_sd      = round(sd(height_s_top,   na.rm = TRUE), 4),
    n              = n(),
    .groups = "drop"
  )

# View it
print(summary_table)

# Export as CSV
#write.csv(summary_table, "trait_summary.csv", row.names = FALSE)
 
# pivot to long format
df_long <- df_traits |>
  pivot_longer(cols = all_of(traits),
               names_to  = "trait",
               values_to = "value") |>
  mutate(trait = factor(trait, levels = traits))
 
# making plot 
traits_plotted <- ggplot(df_long,
            aes(x = age_class, y = value, fill = month, colour = month)) +
 
  # raw points (jittered, dodged to match boxes)
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    size  = 1.6, alpha = 0.45, shape = 16
  ) +
 
  # box plots (dodged, slightly transparent)
  geom_boxplot(
    position     = position_dodge(width = 0.75),
    width        = 0.55,
    outlier.shape = NA,          # hide outlier dots (already shown as points)
    alpha        = 0.35,
    linewidth    = 0.55
  ) +
 
  # mean ± 1 SD error bars
  stat_summary(
    fun.data = mean_se, # standard error around mean
    geom      = "errorbar",
    position  = position_dodge(width = 0.75),
    width     = 0.25,
    linewidth = 0.7
  ) +
 
  # mean dot 
  stat_summary(
    fun      = mean, 
    geom     = "point",
    shape    = 23,               # filled diamond
    size     = 2.5,
    position = position_dodge(width = 0.75)
  ) +
 
  # facet: one panel per trait 
  facet_wrap(~ trait,
             scales   = "free_y",
             labeller = as_labeller(trait_labels),
             ncol     = 3) +
 
  # colour
  scale_fill_manual(  values = c("November" = "#7B2D8B", "January" = "#C589D8")) +
  scale_colour_manual(values = c("November" = "#7B2D8B", "January" = "#C589D8")) +
 
  # labels
  labs(
    title  = "Trait measurements by growth phase and month",
    x      = "Growth phase",
    y      = "Trait Value",
    fill   = "Month",
    colour = "Month"
  ) +
 
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "grey92", colour = "grey60"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 30, hjust = 1)
  )

print(traits_plotted) # must be run in the console

# save fig to output folder 
ggsave(
  here("output", "raw_trait_plot.png"),
  width  = 10,
  height = 8,
  dpi    = 300
)
 
```

## Trait means for each plant individual

Before combining with T50 data, have to reduce the 5 measurements per individual to one mean value. As T50 data only have one value per individual.

```{r}
#| label: means per individual plant  
#| message: false

# collapsing the five (or four) measurements of traits per individual id plant, into the mean of them
trait_data_means <- trait_data |>
  dplyr::group_by(id, month, age_class, plant_nr) |>
  dplyr::summarise(
    leaf_thickness_mean = mean(leaf_thickness_mean, na.rm = TRUE),
    sla = mean(sla, na.rm = TRUE),
    ldmc = mean(ldmc, na.rm = TRUE),
    dry_mass_g = mean(dry_mass_g, na.rm = TRUE),
    height_s_top = unique(height_s_top), # does not need averaging like the other traits 
    .groups = "drop"
  )

```

### Correlation matrix between traits

Only problematic if correlation is above 0.7-0.9.

```{r}
#| label: correlation matrix   
#| message: false

# correlation matrix NOT taking month into account 
trait_cor <- trait_data_means |>
  dplyr::select(leaf_thickness_mean, sla, ldmc, dry_mass_g, height_s_top) |>
  cor(use = "complete.obs")

trait_cor # added height and dry mass, now a strong negative correlation between sla and height, -0.79
# exceeds +/- 0.7 so could potentially be problematic correlations for lm model later
# would then be related to collinearity, check VIF in diagnostics



#correlation matrix taking month into account
trait_cor_by_month <- trait_data_means |>
  dplyr::group_split(month) |>
  setNames(unique(trait_data_means$month)) |>
  lapply(function(df) {
    cor(
      df |> dplyr::select(leaf_thickness_mean, sla, ldmc, dry_mass_g, height_s_top),
      use = "complete.obs"
    )
  })

trait_cor_by_month #  added height and dry mass
# also here strong relashionship between sla and height, both months similar -0.8, -0.84 
# so taller plants = conservative leaves, shorter plants aquisitive leaves, LES
# sla and ldmc moderately negatively correlated around -0.6, as expected in LES.
# LDMC and height moderate positive correlation, taller plants have more conservative leaves
# sla vs dry mass show opposite patterns in nov and jan

```

Traits are not uncorrelated, several relationships detected in line with LES and plant strategy theory.

High correlation between SLA and height, I choose to keep SLA, as I have more litterature on this relationship with T50.

```{r}
#| label: correlation plot figure   
#| message: false
library(corrplot)

names(trait_cor_by_month) <- c("January", "November")

#simple
cor_plot_nov <- corrplot(trait_cor_by_month$November)
cor_plot_jan <- corrplot(trait_cor_by_month$January)

# recode trait names for nicer labels 
trait_labels <- c(
  leaf_thickness_mean = "Leaf Thickness",
  sla                 = "SLA ",
  ldmc                = "LDMC",
  dry_mass_g          = "Dry Mass",
  height_s_top        = "Height"
)
relabel <- function(mat) {
  idx <- match(rownames(mat), names(trait_labels))
  rownames(mat) <- colnames(mat) <- trait_labels[idx]
  mat
}


# better plots for thesis

png("trait_cor_nov_jan.png", width = 1200, height = 550, res = 120)

par(mfrow = c(1, 2))

corrplot(relabel(trait_cor_by_month$November),
         method      = "circle",
         order       = "AOE",
         diag        = TRUE,
         tl.col      = "black",
         addCoef.col = "black",
         number.cex  = 0.8,
         col         = colorRampPalette(c("#67001F", "#FFFFFF", "#4575B4"))(10),
         title       = "November",
         mar         = c(0, 0, 2, 0))

corrplot(relabel(trait_cor_by_month$January),
         method      = "circle",
         order       = "AOE",
         diag        = TRUE,
         tl.col      = "black",
         addCoef.col = "black",
         number.cex  = 0.8,
         col         = colorRampPalette(c("#67001F", "#FFFFFF", "#4575B4"))(10),
         title       = "January",
         mar         = c(0, 0, 2, 0))

dev.off()

# ggsave does not work, corrplot not in ggplot
# reorder traits if time
```

#### Standardising traits

Standardizing units, z-scores by (x - mean(x)) / sd (x). Gives mean = 0 and SD = 1 for each trait. To make coeficcients in model output direclty comparable, so answer which trait has the strongest effect on T50. Will not use this approach after all.

## HOT

## Joining T50 hot data to trait data

```{r}
#| label: joining data frames   
#| message: false

# joining trait means data frame with T50 hot dataframe by id column
trait_T50_hot <- trait_data_means |>
  dplyr::left_join(T50_hot_data, by = "id") |> 
   mutate(month = factor(month, levels = c("November", "January"))) |> 
   mutate(age_class = factor(age_class, levels = c(
    "Newly Burnt", "Pioneer", "Building", "Mature"
  ))) |> 
  na.omit() # remove rows with NA

```

## Exploratory plots: T50 hot against each trait raw value

```{r}
#| label: exploratory plots    
#| message: false

# tried without facteting by month. already know month has an impact, so could mask trait trends 
# faceting by month, as that cold mask trends 

# mean thickness vs T50 hot
thickness_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = leaf_thickness_mean, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "Leaf thickness", y = "T50") +
  theme_minimal()

thickness_vs_T50_hot 
# when not faceted by month: slight increase in T50 hot with increased thickness
# as expected
# when faceted by month: January relative flat relationship, November increased thickness corresponds to decreased T50. not as expected. 

# mean sla vs T50 hot
sla_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = sla, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "SLA", y = "T50") +
  theme_minimal()

sla_vs_T50_hot 
# when NOT faceted by month: slight increase in T50 hot with increased sla,
# not as expected. maybe the very high sla group jan_nb_1 influences this
# when faceted by month: January increased sla corresponds to increased T50, not expected. November flat relationship


# mean ldmc vs T50 hot
ldmc_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = ldmc, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "LDMC", y = "T50") +
  theme_minimal()

ldmc_vs_T50_hot 
# when NOT faceted by month: decrease in T50 hot with increased ldmc, not expected
# when faceted by month: Janury increased ldmc relate to decreased T50, not expected. November increased ldmc relate to slight increase in T50. 

# dry mass vs T50 hot
drymass_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = dry_mass_g, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "Dry mass (g)", y = "T50") +
  theme_minimal()

drymass_vs_T50_hot
# nov slight negative trend, jan positive trend

height_vs_T50_hot <- ggplot(trait_T50_hot, aes(x = height_s_top, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "Plant height", y = "T50") +
  theme_minimal()

height_vs_T50_hot
# slightly negative trend with increasing height 
```

## Linear model for traits vs T50 hot

```{r}
#| label: linear model   
#| message: false

# linear model for T50 response to standardised trait values, no interactions included
lm_traits_T50_hot <- lm(T50 ~ sla + leaf_thickness_mean + ldmc + dry_mass_g + month + age_class, data = trait_T50_hot) 
anova(lm_traits_T50_hot) #look at and report results
summary(lm_traits_T50_hot) # look at and report results
# summed up: thickness, ldmc, month significant predictors


# USE THIS: linear interaction model for T50 response to standardised trait values, no interactions included
lm_int_traits_T50_hot <- lm(T50 ~ (sla + leaf_thickness_mean + ldmc + dry_mass_g) * month * age_class , data = trait_T50_hot) 

car::Anova(lm_int_traits_T50_hot, type = "III") # order does not matter 
summary(lm_int_traits_T50_hot)


AIC(lm_traits_T50_hot, lm_int_traits_T50_hot) #
# AIC favours interaction model
```

## Checking model performance

```{r}
#| label: checking model preformance 
#| message: false

library(performance)
check_model(lm_int_traits_T50_hot)
# interaction model has high VIF for all terms, can inflate parameter uncertainty. Can still use the model. 
```

```{r}
#| label: effect size plot   
#| message: false

# maybe not a good option any more if not standardizing predictors
```

```{r}
#| label: model coefficient table    
#| message: false

tidy(lm_int_traits_T50_hot, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(term, estimate, conf.low, conf.high, statistic, p.value) |>
  gt() |>
  cols_label(term = "Term", estimate = "Estimate",
             conf.low = "95% CI Low", conf.high = "95% CI High",
             statistic = "t", p.value = "p") |>
  gtsave("lm_traits_T50_hot.docx")
```

## COLD

## Joining T50 cold data to trait data

```{r}
#| label: joining data frames   
#| message: false

# joining tarit means data frame with T50 hot dataframe by id column
trait_T50_cold <- trait_data_means |>
  dplyr::left_join(T50_cold_data, by = "id") |> 
   mutate(month = factor(month, levels = c("November", "January"))) |> 
   mutate(age_class = factor(age_class, levels = c(
    "Newly Burnt", "Pioneer", "Building", "Mature"
  ))) |> 
   # now filter out extremely low, low_prop T50 estimates
   filter(T50 >= -35) |> # November building 6
   filter(!is.na(T50))  # only remove rows where T50 itself is NA

```

## Exploratory plots: T50 cold against each trait

```{r}
#| label: exploratory plots    
#| message: false

# faceting by month, as data availability differ

# mean thickness vs T50 cold
thickness_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = leaf_thickness_mean, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "Leaf thickness", y = "T50") +
  theme_minimal()

thickness_vs_T50_cold 
# when not faceted by month: thicker leaves less tolerant to cold. 
# when faceted by month: November, slight increase in thickness correlated to less cold tolerant. January, clearer similar trend NB: very few data points. 

# mean sla vs T50 cold
sla_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = sla, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "SLA", y = "T50") +
  theme_minimal()

sla_vs_T50_cold 
# when NOT faceted by month: higher SLA (more aqusitive) have less cold tolerant leaves. as expected
# when faceted by month: November higher SLA (more aqusitive) have less cold tolerant leaves. as expected. January, opposite pattern, high SLA (aquisitive) correlates to better cold tolerance - NB few data points 

# mean ldmc vs T50 cold
ldmc_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = ldmc, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "LDMC", y = "T50") +
  theme_minimal()

ldmc_vs_T50_cold
# when NOT faceted by month: high LDMC correlates to slightly less cold tolerant leaves. not expected
# when faceted by month: November high LDMC are more cold tolerant. Janyary almost flat relationship - NB few data points. 

#dry mass
drymass_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = dry_mass_g, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "drymass", y = "T50") +
  theme_minimal()

drymass_vs_T50_cold

#height 
height_vs_T50_cold <- ggplot(trait_T50_cold, aes(x = height_s_top, y = T50)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ month) +
  labs(x = "height", y = "T50") +
  theme_minimal()

height_vs_T50_cold
# taller plant, slightly more freezing tolerant
```

## Linear model for traits vs T50 cold

UPDATE 24/4: Confirmed that as we can not calculate viable T50 estimates for January this can not be included in the model. Remove month as predictor in the linear model for cold. No interactions.

```{r}
#| label: linear model cold nov
#| message: false

# filter for november T50 data
trait_T50_cold_nov <- trait_T50_cold |> filter(month == "November")

# linear model for T50 nov response to triats, no interactions included
lm_traits_T50_cold_nov <- lm(T50 ~ sla + leaf_thickness_mean + ldmc + dry_mass_g + age_class, data = trait_T50_cold_nov) 
anova(lm_traits_T50_cold_nov) 
summary(lm_traits_T50_cold_nov)

#interaction model 
lm_int_traits_T50_cold_nov <- lm(T50 ~ (sla + leaf_thickness_mean + ldmc + dry_mass_g) * age_class, data = trait_T50_cold_nov) 

car::Anova(lm_int_traits_T50_cold_nov, type = "III") # order independent 
summary(lm_int_traits_T50_cold_nov) # look at and report results

AIC(lm_traits_T50_cold_nov, lm_int_traits_T50_cold_nov) # 163 vs 161
# AIC favours the interaction model
```

## Checking model performance

```{r}
#| label: checking model preformance 
#| message: false

library(performance)
check_model(lm_int_traits_T50_cold_nov) 
# high VIF for most predictors 
```

```{r}
#| label: cold int mod coef table 
#| message: false

tidy(lm_int_traits_T50_cold_nov, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(term, estimate, conf.low, conf.high, statistic, p.value) |>
  gt() |>
  cols_label(term = "Term", estimate = "Estimate",
             conf.low = "95% CI Low", conf.high = "95% CI High",
             statistic = "t", p.value = "p") |>
  gtsave("lm_traits_T50_cold.docx")
```

**EGU PLOT LDMC COMBINED**

```{r}


library(ggeffects)
library(ggplot2)
library(dplyr)

# ── Partial effects ───────────────────────────────────────────────────────────

p_hot <- ggpredict(lm_traits_T50_hot,
                   terms = c("ldmc", "month")) |>
  as.data.frame() |>
  mutate(
    group     = recode(group, "November" = "Autumn", "January" = "Winter"),
    tolerance = "Heat tolerance"
  )

p_cold <- ggpredict(lm_traits_T50_cold_nov,
                    terms = "ldmc") |>
  as.data.frame() |>
  mutate(
    group     = "Autumn",
    tolerance = "Freezing tolerance"
  )

combined <- bind_rows(p_hot, p_cold)

# ── Raw data ──────────────────────────────────────────────────────────────────

raw_hot <- trait_T50_hot |>
  select(T50, ldmc, month) |>
  mutate(
    month     = recode(month, "November" = "Autumn", "January" = "Winter"),
    tolerance = "Heat tolerance"
  )

raw_cold <- trait_T50_cold_nov |>
  select(T50, ldmc, month) |>
  mutate(
    month     = recode(month, "November" = "Autumn", "January" = "Winter"),
    tolerance = "Freezing tolerance"
  )

raw_all <- bind_rows(raw_hot, raw_cold)

# ── Palette ───────────────────────────────────────────────────────────────────

purple_dark  <- "#7B2D8B"
purple_light <- "#C589D8"

# ── Facet order ───────────────────────────────────────────────────────────────

level_order <- c("Freezing tolerance", "Heat tolerance")  # CHANGED: cold left, heat right

combined$tolerance <- factor(combined$tolerance, levels = level_order)
raw_all$tolerance  <- factor(raw_all$tolerance,  levels = level_order)

# ── Plot ──────────────────────────────────────────────────────────────────────

ggplot() +
  geom_ribbon(data = combined,
              aes(x = x, ymin = conf.low, ymax = conf.high,
                  fill = group),
              alpha = 0.2) +
  geom_line(data = combined,
            aes(x = x, y = predicted, color = group),
            linewidth = 1) +
  geom_point(data = raw_all,
             aes(x = ldmc, y = T50, color = month),
             alpha = 0.5, size = 2) +
  facet_wrap(~ tolerance, scales = "free_y") +
  scale_color_manual(values = c("Autumn" = purple_dark,
                                "Winter" = purple_light)) +
  scale_fill_manual(values  = c("Autumn" = purple_dark,
                                "Winter" = purple_light)) +
  labs(
    x     = expression("LDMC (mg g"^-1*")"),
    y     = "T50 (\u00b0C)",
    color = "Season",
    fill  = "Season"
  ) +
  guides(color = guide_legend(title = "Season"),  # MOVED outside theme()
         fill  = guide_legend(title = "Season")) +
  theme_bw() +
  theme(
    strip.text      = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    legend.title    = element_text(face = "bold"),
     legend.margin   = margin(t = -10) 
  )


ggsave(
  here("output", "ldmc_vs_T50_hot_cold_deg.png"),
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r}
# adding colour and lines for age classes

# ── Partial effects per age class ─────────────────────────────────────────────
p_hot_age <- ggpredict(lm_traits_T50_hot,
                       terms = c("ldmc", "month", "age_class")) |>
  as.data.frame() |>
  rename(age_class = facet) |> 
  mutate(
    group     = recode(group, "November" = "Autumn", "January" = "Winter"),
    tolerance = "Heat tolerance"
  ) |>
  select(-any_of("panel"))

p_cold_age <- ggpredict(lm_traits_T50_cold_nov,
                        terms = c("ldmc", "age_class")) |>
  as.data.frame() |>
  rename(age_class = group) |>  
  mutate(
    group     = "Autumn",
    tolerance = "Freezing tolerance"
  ) 

combined_age <- bind_rows(p_hot_age, p_cold_age) |>
  mutate(
    tolerance = factor(tolerance, levels = level_order),
    age_class = factor(age_class, levels = c(
      "Newly Burnt", "Pioneer", "Building", "Mature"
    ))
  )

# ── Colour palette (same as previous script) ──────────────────────────────────
age_colours <- c(
  "Newly Burnt" = "#D55E00",
  "Pioneer"     = "#E69F00",
  "Building"    = "#56B4E9",
  "Mature"      = "#009E73"
)

# ── Raw data ───────────────────────────────────────────────────────────────────
raw_hot <- trait_T50_hot |>
  select(T50, ldmc, month, age_class) |>
  mutate(
    month     = recode(month, "November" = "Autumn", "January" = "Winter"),
    tolerance = "Heat tolerance"
  )
raw_cold <- trait_T50_cold_nov |>
  select(T50, ldmc, month, age_class) |>
  mutate(
    month     = recode(month, "November" = "Autumn", "January" = "Winter"),
    tolerance = "Freezing tolerance"
  )
raw_all <- bind_rows(raw_hot, raw_cold) |>
  mutate(
    age_class = factor(age_class, levels = c(
      "Newly Burnt", "Pioneer", "Building", "Mature"
    ))
  )
```

```{r}
# ── Plot ───────────────────────────────────────────────────────────────────────
#install.packages("ggnewscale")
library(ggnewscale)


#---- line for each age class

ggplot() +
  geom_ribbon(data = combined,
              aes(x = x, ymin = conf.low, ymax = conf.high,
                  fill = group),
              alpha = 0.2) +
  geom_line(data = combined,
            aes(x = x, y = predicted, color = group, linetype = group),
            linewidth = 1) +
  scale_color_manual( 
                     name = "Season",
                     values = c("Autumn" = purple_dark,
                                "Winter" = purple_light)) +
  scale_fill_manual(
                    name = "Season",
                    values = c("Autumn" = purple_dark,
                               "Winter" = purple_light)) +
  scale_linetype_manual(
                        name = "Season", 
            values = c("Autumn" = "solid",
                       "Winter" = "dashed")) +
  new_scale_color() +
  new_scale_fill() +

  # ── Per-age-class predicted lines ──────────────────────────────────────────
 
  geom_ribbon(data = combined_age,
              aes(x = x, ymin = conf.low, ymax = conf.high,
                  fill = age_class,
                  group = interaction(age_class, group)),
              alpha = 0.15,
              show.legend = FALSE) +
  
geom_line(data = combined_age,
            aes(x = x, y = predicted,
                colour = age_class,
                group  = interaction(age_class, group),
                linetype  = group 
                ),
            linewidth = 0.8,
            show.legend = FALSE) +
  
  
  geom_point(data = raw_all,
             aes(x = ldmc, y = T50,
                 colour = age_class),
             alpha = 0.8, size = 2.5) +
  scale_color_manual(name   = "Successional stage",
                     values = age_colours) +
  scale_fill_manual(values  = age_colours,
                    guide   = "none") +

  facet_wrap(~ tolerance, scales = "free_y") +
  labs(
    x = expression("LDMC (mg g"^-1*")"),
    y = "T50 (C)"
  ) +
  guides(
    fill   = guide_legend(
      title = "Season", nrow = 1,
      override.aes = list(shape = 21, colour = "white")
    ),
    colour = guide_legend(title = "Successional stage", nrow = 1)
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 14),
    legend.position  = "bottom",
    legend.box       = "vertical",
    legend.direction = "horizontal",
    legend.title     = element_text(face = "bold"),
    legend.margin    = margin(t = -10)
  )


ggsave(
  here("output", "ldmc_vs_T50_hot_cold_colour_lines.png"),
  width = 8,
  height = 6,
  dpi = 300
)  
```

```         
```
