---
title: "Traits"
format: html
editor: visual
---

## Loading library

```{r}
#| label: loading library
#| message: false

library(tidyverse)
library(here)
```

## Data import

```{r}
#| label: data import 
#| message: false

clean_data <- read.csv(here("clean_data", "DURIN_Clean_Seasonal_Thermal_Tolerance_Bente_Msc.csv")) 
  filter

view(clean_data) # check import 
```

## Mean table and exploratory plot

find out if I want leaf are in here as well

```{r}
#| label: exploratory plot 
#| message: false

# make trait object
traits <- c("leaf_thickness_mean", "sla", "ldmc")
 
# tidy labels for facet strips
trait_labels <- c(
  leaf_thickness_mean = "Leaf Thickness (mm)",
  sla                 = "SLA (cm/g)",
  ldmc                = "LDMC (mg/g)"
)
 
df_traits <- clean_data |>
  filter(proj_2025 == "Traits") |>
  select(plant_nr, age_class, month, all_of(traits)) |>   # keep relevant cols
  mutate(
    month     = factor(month,     levels = c("November", "January"), # adjusting levels for fig
                       labels = c("November", "January")),
    age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer", 
                                             "Building", "Mature"))                           
  )


# want a summary table of means of trait values  pr group
summary_table <- df_traits |>
  group_by(age_class, month) |>
  summarise(
    thickness_mean = round(mean(leaf_thickness_mean, na.rm = TRUE), 4),
    thickness_sd   = round(sd(leaf_thickness_mean,   na.rm = TRUE), 4),
    sla_mean       = round(mean(sla,                 na.rm = TRUE), 4),
    sla_sd         = round(sd(sla,                   na.rm = TRUE), 4),
    ldmc_mean      = round(mean(ldmc,                na.rm = TRUE), 4),
    ldmc_sd        = round(sd(ldmc,                  na.rm = TRUE), 4),
    n              = n(),
    .groups = "drop"
  )

# View it
print(summary_table)

# Export as CSV
#write.csv(summary_table, "trait_summary.csv", row.names = FALSE)
 
# pivot to long format
df_long <- df_traits |>
  pivot_longer(cols = all_of(traits),
               names_to  = "trait",
               values_to = "value") |>
  mutate(trait = factor(trait, levels = traits))
 
# making plot 
traits_plotted <- ggplot(df_long,
            aes(x = age_class, y = value, fill = month, colour = month)) +
 
  # raw points (jittered, dodged to match boxes)
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    size  = 1.6, alpha = 0.45, shape = 16
  ) +
 
  # box plots (dodged, slightly transparent)
  geom_boxplot(
    position     = position_dodge(width = 0.75),
    width        = 0.55,
    outlier.shape = NA,          # hide outlier dots (already shown as points)
    alpha        = 0.35,
    linewidth    = 0.55
  ) +
 
  # mean ± 1 SD error bars
  stat_summary(
    fun.data = mean_se, # standard error around mean
    geom      = "errorbar",
    position  = position_dodge(width = 0.75),
    width     = 0.25,
    linewidth = 0.7
  ) +
 
  # mean dot 
  stat_summary(
    fun      = mean, 
    geom     = "point",
    shape    = 23,               # filled diamond
    size     = 2.5,
    position = position_dodge(width = 0.75)
  ) +
 
  # facet: one panel per trait 
  facet_wrap(~ trait,
             scales   = "free_y",
             labeller = as_labeller(trait_labels),
             ncol     = 3) +
 
  # colour
  scale_fill_manual(  values = c("November" = "#E07B4F", "January" = "#4E9EC2")) +
  scale_colour_manual(values = c("November" = "#E07B4F", "January" = "#4E9EC2")) +
 
  # labels
  labs(
    title  = "Leaf Trait Measurements by Age Class and Month",
    x      = "Age Class",
    y      = "Trait Value",
    fill   = "Month",
    colour = "Month"
  ) +
 
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "grey92", colour = "grey60"),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "top",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 30, hjust = 1)
  )

print(traits_plotted) # must be run in the console

# add ggsave when happy
 
```

Leaf thickness big spread, no clear trends visually. Maybe generally somewhat thicker in January, except for pio.

SLA clearly higher for Newly Burnt than for the other age classes. Looks relative stable between seasons.

Looks like LDMC is clearly higher in November than in January. Also increasing with age potentially.

Have to test if there is actually significant difference between groups. Want to compare using a LMM, must check if normality assumption is met first.

## Check if normality assumption is met

```{r}

df_long |>
  dplyr::filter(trait %in% c("leaf_thickness_mean", "sla", "ldmc")) |>
  group_by(trait, age_class, month) |>
  summarise(shapiro_p = shapiro.test(value)$p.value)
```

## Log-transformation of data to meet normality assumption

```{r}

# Leaf thickness — try log transform to fix the Pioneer November group
df_traits <- df_traits |>
  mutate(
    leaf_thickness_log = log(leaf_thickness_mean),
    ldmc_log           = log(ldmc),
    sla_log            = log(sla)
  )
```

## Model fitting on log-transformed data

```{r}

library(lme4)
library(lmerTest)

conflicted::conflicts_prefer(lmerTest::lmer)

# Fit models — using log-transformed versions
model_thickness <- lmer(leaf_thickness_log ~ age_class * month + (1 | plant_nr), 
                        data = df_traits)
model_sla       <- lmer(sla_log ~ age_class * month + (1 | plant_nr), 
                        data = df_traits)
model_ldmc      <- lmer(ldmc_log ~ age_class * month + (1 | plant_nr), 
                        data = df_traits)
```

## Check the p-values

```{r}


# Get ANOVA-style p-values for each term
# This is NOT actually an ANOVA, just gives readable output on the same format

anova(model_thickness)
anova(model_sla)
anova(model_ldmc)
```

clearly significant for age class and month seperatly for thickness and sla, and ldmc less

interaction term less so for ldmc

## Checking residuals

Using residuals vs fitted - if normality then points should be scattered randomly

Using QQ-plot - points should be on a straight line, heavy tails can indicate problems

```{r}

# Leaf thickness
plot(model_thickness, main = "Leaf Thickness - Residuals vs Fitted") # looks good
qqnorm(resid(model_thickness), main = "Leaf Thickness - QQ Plot") #looks good
qqline(resid(model_thickness))

# SLA
plot(model_sla, main = "SLA - Residuals vs Fitted") #looks good
qqnorm(resid(model_sla), main = "SLA - QQ Plot") #looks good
qqline(resid(model_sla))

# LDMC
plot(model_ldmc, main = "LDMC - Residuals vs Fitted") #looks good
qqnorm(resid(model_ldmc), main = "LDMC - QQ Plot") # has heavy tails, but ok
qqline(resid(model_ldmc))
```

## Post-hoc tests

Pairwise comparisons within month, within age class.

```{r}
# comparing pairwise 

library(emmeans)

# Compare age classes within each month
emmeans(model_thickness, pairwise ~ age_class | month, adjust = "tukey")
emmeans(model_sla,       pairwise ~ age_class | month, adjust = "tukey")
emmeans(model_ldmc,      pairwise ~ age_class | month, adjust = "tukey")

# Compare months within each age class
emmeans(model_thickness, pairwise ~ month | age_class, adjust = "tukey")
emmeans(model_sla,       pairwise ~ month | age_class, adjust = "tukey")
emmeans(model_ldmc,      pairwise ~ month | age_class, adjust = "tukey")
```

Big output, need to look into later.
